
---
title: "Glasgow Clustering"
author: "Guanyu"
date: "`r Sys.Date()`"
output:
  pdf_document:
    latex_engine: xelatex
    number_sections: true
    toc: true
  html_document:
    toc: true
    df_print: paged
params:
  data_dir: simd2020_withgeog
  csv_file: simd2020_withinds.csv
  shp_file: sc_dz_11.shp
  target_board: Glasgow
  output_dir: outputs/glasgow
  seed: 123
  missing_threshold: 10
  k_grid: !r 1:10
  chosen_k_raw: 2
  chosen_k_pc: 2
  cmeans_m_grid: !r seq(1.1, 2.5, by = 0.1)
  cmeans_ref_m_raw: 2
  cmeans_ref_m_pc: 2
  cmeans_main_ambiguity_threshold: 0.6
  cmeans_secondary_ambiguity_threshold: 0.7
  vine_core_membership_threshold: 0.8
  use_cmeans: true
  use_mclust: true
---

# How to use this document

This file is meant to be a **reproducible template** for a joint-board-area clustering study.

To run the same workflow for another joint board area, update:

- `params$target_board`
- `params$chosen_k_raw`
- `params$chosen_k_pc`
- `params$cmeans_ref_m_raw`
- `params$cmeans_ref_m_pc`
- `params$k_grid`
- `params$cmeans_m_grid`
- `params$cmeans_main_ambiguity_threshold`
- `params$cmeans_secondary_ambiguity_threshold`
- `params$output_dir`

The workflow is:

1. **Data preprocessing**
   - missingness
   - descriptive statistics
   - distributions
   - correlation structure
   - domain-wise PCA
2. **Clustering**
   - K-means (headline partitioning method)
   - Hierarchical clustering (hard-partition benchmark)
   - C-means (sensitivity / uncertainty layer)
   - Finite mixture models (secondary model-based comparator)
3. **Comparison**
   - cross-tables for selected clustering results
   - combined diagnostic plots
   - cluster maps
   - difference maps

The design follows the same analysis logic already used in your exploratory document: missingness review, descriptive statistics, domain-grouped distributions, correlation heatmaps, and domain-wise PCA before clustering.

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  results = "markup",
  fig.align = "center",
  fig.width = 10,
  fig.height = 7,
  out.width = "100%",
  dev = "pdf"
)
```

```{r packages, include=FALSE}
required_pkgs <- c(
  "tidyverse",
  "sf",
  "cluster",
  "factoextra",
  "patchwork",
  "e1071",
  "mclust",
  "aricode",
  "salso",
  "knitr",
  "kableExtra",
  "reshape2",
  "gtools"
)

missing_pkgs <- required_pkgs[
  !vapply(required_pkgs, requireNamespace, logical(1), quietly = TRUE)
]

if (length(missing_pkgs) > 0) {
  stop(
    paste0(
      "Missing required packages: ",
      paste(missing_pkgs, collapse = ", "),
      "\nInstall them first, e.g.:\ninstall.packages(c(",
      paste(sprintf("\"%s\"", missing_pkgs), collapse = ", "),
      "))"
    ),
    call. = FALSE
  )
}

suppressPackageStartupMessages({
  library(tidyverse)
  library(sf)
  library(cluster)
  library(factoextra)
  library(patchwork)
  library(e1071)
  library(mclust)
  library(aricode)
  library(salso)
  library(knitr)
  library(kableExtra)
  library(reshape2)
})

set.seed(params$seed)

caption_safe <- function(x) {
  gsub("_", " ", x, fixed = TRUE)
}

board_slug <- function(x) {
  x %>%
    stringr::str_to_lower() %>%
    stringr::str_replace_all("[^a-z0-9]+", "_") %>%
    stringr::str_replace_all("^_+|_+$", "")
}

output_prefix <- board_slug(params$target_board)
```

# Shared objects

```{r shared-objects}
`%||%` <- function(x, y) if (is.null(x)) y else x

joint_board_lookup <- tibble::tribble(
  ~Council_area,                          ~Joint_Board_area,
  "East Ayrshire",                       "Ayrshire",
  "North Ayrshire",                      "Ayrshire",
  "South Ayrshire",                      "Ayrshire",
  "Scottish Borders",                    "Borders",
  "Clackmannanshire",                    "Central Scotland",
  "Falkirk",                             "Central Scotland",
  "Stirling",                            "Central Scotland",
  "Dumfries and Galloway",               "Dumfries and Galloway",
  "Argyll and Bute",                     "Dunbartonshire and Argyll & Bute",
  "East Dunbartonshire",                 "Dunbartonshire and Argyll & Bute",
  "West Dunbartonshire",                 "Dunbartonshire and Argyll & Bute",
  "Fife",                                "Fife",
  "Aberdeen City",                       "Grampian",
  "Aberdeenshire",                       "Grampian",
  "Moray",                               "Grampian",
  "Glasgow City",                        "Glasgow",
  "Highland",                            "Highlands and Islands",
  "Na h-Eileanan an Iar",                "Highlands and Islands",
  "North Lanarkshire",                   "Lanarkshire",
  "South Lanarkshire",                   "Lanarkshire",
  "East Lothian",                        "Lothian",
  "City of Edinburgh",                   "Lothian",
  "Midlothian",                          "Lothian",
  "West Lothian",                        "Lothian",
  "Orkney Islands",                      "Orkney and Shetland",
  "Shetland Islands",                    "Orkney and Shetland",
  "East Renfrewshire",                   "Renfrewshire",
  "Inverclyde",                          "Renfrewshire",
  "Renfrewshire",                        "Renfrewshire",
  "Angus",                               "Tayside",
  "Dundee City",                         "Tayside",
  "Perth and Kinross",                   "Tayside"
)

indicator_cols <- c(
  "income_rate", "income_count",
  "employment_rate", "employment_count",
  "CIF", "ALCOHOL", "DRUG", "SMR", "DEPRESS",
  "LBWT", "EMERG", "Attendance", "Attainment",
  "no_qualifications", "not_participating", "University",
  "crime_count", "crime_rate",
  "overcrowded_count", "nocentralheating_count",
  "overcrowded_rate", "nocentralheating_rate",
  "drive_petrol", "drive_GP", "drive_post",
  "drive_primary", "drive_retail", "drive_secondary",
  "PT_GP", "PT_post", "PT_retail", "broadband"
)

indicator_domains <- tibble::tribble(
  ~Indicator,                ~Domain,
  "income_rate",             "Income",
  "income_count",            "Income",
  "employment_rate",         "Employment",
  "employment_count",        "Employment",
  "CIF",                     "Health",
  "ALCOHOL",                 "Health",
  "DRUG",                    "Health",
  "SMR",                     "Health",
  "DEPRESS",                 "Health",
  "LBWT",                    "Health",
  "EMERG",                   "Health",
  "Attendance",              "Education",
  "Attainment",              "Education",
  "no_qualifications",       "Education",
  "not_participating",       "Education",
  "University",              "Education",
  "drive_petrol",            "Access",
  "drive_GP",                "Access",
  "drive_post",              "Access",
  "drive_primary",           "Access",
  "drive_retail",            "Access",
  "drive_secondary",         "Access",
  "PT_GP",                   "Access",
  "PT_post",                 "Access",
  "PT_retail",               "Access",
  "broadband",               "Access",
  "crime_count",             "Crime",
  "crime_rate",              "Crime",
  "overcrowded_count",       "Housing",
  "nocentralheating_count",  "Housing",
  "overcrowded_rate",        "Housing",
  "nocentralheating_rate",   "Housing"
)

domain_order <- c("Income", "Employment", "Health", "Education", "Access", "Crime", "Housing")

group_order <- c("Rates", "Health", "Access", "Education", "Other")

core_domains <- c("Income", "Employment", "Crime", "Housing")
core_rate_vars <- c(
  "income_rate",
  "employment_rate",
  "crime_rate",
  "overcrowded_rate",
  "nocentralheating_rate"
)

cluster_palette <- c(
  "1" = "#1f78b4",
  "2" = "#e31a1c",
  "3" = "#33a02c",
  "4" = "#ff7f00",
  "5" = "#6a3d9a",
  "6" = "#a6cee3",
  "7" = "#b2df8a"
)

domain_palette <- c(
  "Income" = "#1f77b4",
  "Employment" = "#2ca02c",
  "Health" = "#d62728",
  "Education" = "#9467bd",
  "Access" = "#ff7f0e",
  "Crime" = "#17becf",
  "Housing" = "#8c564b"
)
```

# Helper functions

```{r helpers}
assert_file_exists <- function(path) {
  if (!file.exists(path)) stop("File not found: ", path)
  invisible(path)
}

clean_percent_numeric <- function(df, id_col = "DataZone") {
  df %>%
    mutate(
      across(
        -all_of(id_col),
        ~ if (is.character(.x)) {
          suppressWarnings(as.numeric(stringr::str_remove(.x, "%")))
        } else {
          .x
        }
      )
    )
}

scale_df <- function(df, id_col = "DataZone") {
  out <- df
  out[names(out) != id_col] <- scale(out[names(out) != id_col]) %>% as.data.frame()
  out
}

make_missing_summary <- function(df, id_col = "DataZone") {
  df %>%
    summarise(
      across(
        -all_of(id_col),
        list(
          Missing_Count = ~ sum(is.na(.)),
          Percent_Missing = ~ mean(is.na(.)) * 100
        ),
        .names = "{.col}.{.fn}"
      )
    ) %>%
    pivot_longer(
      everything(),
      names_to = c("Indicator", ".value"),
      names_sep = "\\."
    ) %>%
    arrange(desc(Percent_Missing), Indicator)
}

describe_var <- function(x) {
  x <- x[!is.na(x)]
  n <- length(x)

  if (n < 3) {
    return(tibble(
      n = n,
      mean = NA_real_,
      median = NA_real_,
      sd = NA_real_,
      skewness = NA_real_,
      kurtosis = NA_real_,
      Normality_Test = NA_character_,
      p_value = NA_real_
    ))
  }

  p_val <- tryCatch(stats::shapiro.test(x)$p.value, error = function(e) NA_real_)

  tibble(
    n = n,
    mean = mean(x),
    median = median(x),
    sd = sd(x),
    skewness = e1071::skewness(x, type = 2),
    kurtosis = e1071::kurtosis(x, type = 2),
    Normality_Test = "Shapiro-Wilk",
    p_value = p_val
  )
}

format_stats_table <- function(df, id_col = "DataZone") {
  df %>%
    select(-all_of(id_col)) %>%
    map_dfr(describe_var, .id = "Indicator") %>%
    mutate(
      across(c(mean, median, sd, skewness, kurtosis), ~ round(.x, 2)),
      p_value = ifelse(
        is.na(p_value), 
        NA_character_,
        ifelse(p_value < 0.0001, "<0.0001", sprintf("%.4f", p_value))
      )
    )
}

first_existing_col <- function(df, candidates) {
  candidates <- unique(candidates[!is.na(candidates) & nzchar(candidates)])
  hit <- intersect(candidates, names(df))

  if (length(hit) == 0) {
    return(NULL)
  }

  hit[[1]]
}

best_metric_index <- function(x) {
  x <- suppressWarnings(as.numeric(x))

  if (length(x) == 0 || all(!is.finite(x))) {
    return(NA_integer_)
  }

  which.max(replace(x, !is.finite(x), -Inf))
}

best_metric_k <- function(diag_tbl, metric_col = "silhouette", k_col = "k") {
  if (is.null(diag_tbl) || nrow(diag_tbl) == 0 || !all(c(metric_col, k_col) %in% names(diag_tbl))) {
    return(NA_integer_)
  }

  idx <- best_metric_index(diag_tbl[[metric_col]])
  if (is.na(idx)) return(NA_integer_)

  as.integer(diag_tbl[[k_col]][idx])
}

best_metric_value <- function(diag_tbl, metric_col = "silhouette") {
  if (is.null(diag_tbl) || nrow(diag_tbl) == 0 || !(metric_col %in% names(diag_tbl))) {
    return(NA_real_)
  }

  vals <- suppressWarnings(as.numeric(diag_tbl[[metric_col]]))
  if (length(vals) == 0 || all(!is.finite(vals))) {
    return(NA_real_)
  }

  max(vals[is.finite(vals)])
}

domain_map_for_features <- function(feature_names) {
  pca_map <- tibble(
    Indicator = feature_names[stringr::str_detect(feature_names, "_PC[0-9]+$")],
    Domain = stringr::str_extract(feature_names[stringr::str_detect(feature_names, "_PC[0-9]+$")], "^[A-Za-z]+")
  )

  bind_rows(indicator_domains, pca_map) %>%
    filter(Indicator %in% feature_names) %>%
    distinct() %>%
    mutate(Domain = factor(Domain, levels = domain_order))
}

group_map_for_features <- function(feature_names,
                                   group_levels = get0(
                                     "group_order",
                                     ifnotfound = c("Rates", "Health", "Access", "Education", "Other"),
                                     inherits = TRUE
                                   )) {
  domain_map_for_features(feature_names) %>%
    mutate(
      Group = case_when(
        Domain %in% c("Income", "Employment", "Crime", "Housing") ~ "Rates",
        as.character(Domain) == "Health" ~ "Health",
        as.character(Domain) == "Access" ~ "Access",
        as.character(Domain) == "Education" ~ "Education",
        TRUE ~ "Other"
      ),
      Group = factor(Group, levels = group_levels)
    )
}

feature_representation_summary <- function(raw_features, pc_features) {
  tibble(Domain = factor(domain_order, levels = domain_order)) %>%
    mutate(
      raw_representation = purrr::map_chr(Domain, function(d) {
        vals <- domain_map_for_features(raw_features) %>%
          filter(as.character(Domain) == as.character(d)) %>%
          pull(Indicator)
        if (length(vals) == 0) return("—")
        paste(vals, collapse = ", ")
      }),
      pc_representation = purrr::map_chr(Domain, function(d) {
        vals <- domain_map_for_features(pc_features) %>%
          filter(as.character(Domain) == as.character(d)) %>%
          pull(Indicator)
        if (length(vals) == 0) return("—")
        paste(vals, collapse = ", ")
      })
    ) %>%
    rename(
      `Retained raw indicators` = raw_representation,
      `PC-based representation` = pc_representation
    )
}

plot_domain_distributions <- function(df,
                                      id_col = "DataZone",
                                      title_text = "Distribution plot",
                                      geom_type = c("histogram", "density"),
                                      ncol = 4,
                                      bins = 30,
                                      by = c("domain", "group"),
                                      group_levels = get0(
                                        "group_order",
                                        ifnotfound = c("Rates", "Health", "Access", "Education", "Other"),
                                        inherits = TRUE
                                      )) {
  geom_type <- match.arg(geom_type)
  by <- match.arg(by)

  feature_names <- setdiff(names(df), id_col)

  if (by == "group") {
    meta_map <- group_map_for_features(feature_names, group_levels = group_levels) %>%
      rename(Colour_Group = Group)
    fill_values <- c(
      "Rates" = "#4C78A8",
      "Health" = "#E45756",
      "Access" = "#F58518",
      "Education" = "#8F63B8",
      "Other" = "grey70"
    )
    fill_lab <- "Group"
    order_levels <- meta_map %>% arrange(Colour_Group, Domain, Indicator) %>% pull(Indicator)
  } else {
    meta_map <- domain_map_for_features(feature_names) %>%
      rename(Colour_Group = Domain)
    fill_values <- domain_palette
    fill_lab <- "Domain"
    order_levels <- meta_map %>% arrange(Colour_Group, Indicator) %>% pull(Indicator)
  }

  long_df <- df %>%
    pivot_longer(-all_of(id_col), names_to = "Indicator", values_to = "Value") %>%
    left_join(meta_map, by = "Indicator") %>%
    mutate(
      Colour_Group = factor(Colour_Group, levels = names(fill_values)),
      Indicator = factor(Indicator, levels = order_levels)
    )

  p <- ggplot(long_df, aes(x = Value, fill = Colour_Group, colour = Colour_Group))

  if (geom_type == "histogram") {
    p <- p +
      geom_histogram(aes(y = after_stat(density)), bins = bins, alpha = 0.70, linewidth = 0.25)
  } else {
    p <- p +
      geom_density(alpha = 0.30, linewidth = 0.8)
  }

  p +
    facet_wrap(~ Indicator, scales = "free", ncol = ncol) +
    scale_fill_manual(values = fill_values, drop = FALSE) +
    scale_colour_manual(values = fill_values, drop = FALSE, guide = "none") +
    labs(title = title_text, x = "Value", y = "Density", fill = fill_lab) +
    theme_minimal(base_size = 12) +
    theme(
      plot.title = element_text(face = "bold", hjust = 0.5),
      strip.text = element_text(face = "bold"),
      legend.position = "right"
    )
}

plot_correlation_heatmap <- function(df,
                                     id_col = "DataZone",
                                     title_text = "Correlation heatmap") {
  num_df <- df %>% select(-all_of(id_col))
  corr_matrix <- cor(num_df, use = "pairwise.complete.obs")

  domain_map <- domain_map_for_features(colnames(num_df))
  ordered_features <- domain_map %>%
    arrange(Domain, Indicator) %>%
    pull(Indicator) %>%
    intersect(colnames(corr_matrix))

  corr_long <- reshape2::melt(corr_matrix, varnames = c("Var1", "Var2"), value.name = "Correlation") %>%
    mutate(
      Var1 = factor(Var1, levels = ordered_features),
      Var2 = factor(Var2, levels = ordered_features)
    ) %>%
    drop_na(Var1, Var2)

  ggplot(corr_long, aes(x = Var1, y = Var2, fill = Correlation)) +
    geom_tile(color = "white", linewidth = 0.2) +
    scale_fill_gradient2(
      low = "#2166ac",
      mid = "white",
      high = "#b2182b",
      midpoint = 0,
      limits = c(-1, 1),
      name = "Correlation"
    ) +
    coord_fixed() +
    labs(title = title_text, x = "Indicator", y = "Indicator") +
    theme_minimal(base_size = 12) +
    theme(
      plot.title = element_text(face = "bold", hjust = 0.5),
      panel.grid = element_blank(),
      axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
      axis.text.y = element_text(size = 8)
    )
}

run_domain_pca <- function(df, domain_map, domains, variance_cutoff = 0.10) {
  purrr::map(domains, function(domain) {
    vars <- domain_map %>%
      filter(Domain == domain, Indicator %in% names(df)) %>%
      pull(Indicator)

    if (length(vars) == 0) {
      return(list(
        domain = domain,
        pca = NULL,
        variance = NULL,
        retained_scores = NULL,
        loadings = NULL
      ))
    }

    pca <- prcomp(df[, vars, drop = FALSE], center = TRUE, scale. = TRUE)
    var_explained <- (pca$sdev^2) / sum(pca$sdev^2)
    keep_idx <- which(var_explained > variance_cutoff)

    variance_tbl <- tibble(
      Domain = domain,
      PC = paste0("PC", seq_along(var_explained)),
      Variance = var_explained,
      Retained = seq_along(var_explained) %in% keep_idx
    )

    retained_scores <- NULL
    loadings_tbl <- NULL

    if (length(keep_idx) > 0) {
      retained_scores <- as_tibble(pca$x[, keep_idx, drop = FALSE])
      names(retained_scores) <- paste0(domain, "_PC", keep_idx)

      loadings_tbl <- as_tibble(pca$rotation[, keep_idx, drop = FALSE], rownames = "Indicator")
      names(loadings_tbl)[-1] <- paste0(domain, "_PC", keep_idx)
    }

    list(
      domain = domain,
      pca = pca,
      variance = variance_tbl,
      retained_scores = retained_scores,
      loadings = loadings_tbl
    )
  }) %>%
    rlang::set_names(domains)
}

plot_domain_variance <- function(var_df, title_text) {
  pc_levels <- var_df %>%
    mutate(.pc_num = readr::parse_number(PC)) %>%
    arrange(.pc_num) %>%
    pull(PC)

  var_df <- var_df %>%
    mutate(PC = factor(PC, levels = pc_levels))

  ggplot(var_df, aes(x = PC, y = Variance, fill = Retained)) +
    geom_col() +
    geom_text(aes(label = sprintf("%.1f%%", Variance * 100)), vjust = -0.3, size = 3) +
    scale_fill_manual(values = c("TRUE" = "#1f78b4", "FALSE" = "grey75"), guide = "none") +
    labs(title = title_text, x = NULL, y = "Variance explained") +
    theme_minimal(base_size = 11) +
    theme(
      plot.title = element_text(face = "bold", hjust = 0.5),
      axis.text.x = element_text(angle = 45, hjust = 1)
    )
}

feature_set_dictionary <- function(raw_features, pc_features) {
  raw_tbl <- tibble(Feature_Set = "raw", Indicator = raw_features) %>%
    left_join(domain_map_for_features(raw_features), by = "Indicator")

  pc_tbl <- tibble(Feature_Set = "pc", Indicator = pc_features) %>%
    left_join(domain_map_for_features(pc_features), by = "Indicator")

  bind_rows(raw_tbl, pc_tbl) %>%
    mutate(Domain = factor(Domain, levels = domain_order)) %>%
    arrange(Feature_Set, Domain, Indicator)
}

fit_kmeans_labels <- function(x, k, seed = 123, nstart = 50) {
  set.seed(seed)
  stats::kmeans(x, centers = k, nstart = nstart)$cluster
}

fit_hclust_labels <- function(x, k) {
  hc <- hclust(dist(x), method = "ward.D2")
  cutree(hc, k = k)
}

fit_cmeans_model <- function(x, k, m = 2, seed = 123) {
  set.seed(seed)
  e1071::cmeans(x, centers = k, m = m)
}

fit_cmeans_labels <- function(x, k, m = 2, seed = 123) {
  fit <- fit_cmeans_model(x, k = k, m = m, seed = seed)
  max.col(fit$membership)
}

fit_mclust_model <- function(x, k, seed = 123) {
  set.seed(seed)
  mclust::Mclust(x, G = k)
}

fit_mclust_labels <- function(x, k, seed = 123) {
  fit_mclust_model(x, k = k, seed = seed)$classification
}

partition_wss <- function(x, labels) {
  labels <- as.integer(as.factor(labels))
  split_idx <- split(seq_len(nrow(x)), labels)

  sum(purrr::map_dbl(split_idx, function(idx) {
    block <- x[idx, , drop = FALSE]
    center <- colMeans(block)
    sum(rowSums((sweep(block, 2, center, FUN = "-"))^2))
  }))
}

partition_silhouette <- function(x, labels) {
  labels <- as.integer(as.factor(labels))
  if (length(unique(labels)) < 2) return(NA_real_)
  sil <- cluster::silhouette(labels, dist(x))
  mean(sil[, "sil_width"])
}

cmeans_validity <- function(fit) {
  U <- fit$membership
  tibble(
    partition_coefficient = mean(rowSums(U^2)),
    partition_entropy = -mean(rowSums(U * log(pmax(U, 1e-12))))
  )
}

cmeans_membership_summary <- function(fit,
                                      main_threshold = 0.6,
                                      secondary_threshold = 0.7) {
  max_membership <- apply(fit$membership, 1, max)

  tibble(
    mean_max_membership = mean(max_membership),
    prop_max_below_main = mean(max_membership < main_threshold),
    prop_max_below_secondary = mean(max_membership < secondary_threshold)
  )
}

method_diagnostics <- function(x, k_grid, method, seed = 123, m = 2) {
  x_mat <- as.matrix(x)

  purrr::map_dfr(k_grid, function(k) {
    if (k == 1) {
      out <- tibble(
        method = method,
        k = 1,
        wss = partition_wss(x_mat, rep(1L, nrow(x_mat))),
        silhouette = 0
      )

      if (method == "cmeans") {
        out <- out %>%
          bind_cols(tibble(
            partition_coefficient = NA_real_,
            partition_entropy = NA_real_,
            m = m
          ))
      } else if (method == "mclust") {
        out <- out %>%
          bind_cols(tibble(BIC = NA_real_, ICL = NA_real_))
      }

      return(out)
    }

    if (method == "kmeans") {
      labels <- fit_kmeans_labels(x_mat, k = k, seed = seed)
      out <- tibble(
        method = method,
        k = k,
        wss = partition_wss(x_mat, labels),
        silhouette = partition_silhouette(x_mat, labels)
      )
    } else if (method == "hierarchical") {
      labels <- fit_hclust_labels(x_mat, k = k)
      out <- tibble(
        method = method,
        k = k,
        wss = partition_wss(x_mat, labels),
        silhouette = partition_silhouette(x_mat, labels)
      )
    } else if (method == "cmeans") {
      fit <- fit_cmeans_model(x_mat, k = k, m = m, seed = seed)
      labels <- max.col(fit$membership)
      out <- tibble(
        method = method,
        k = k,
        wss = partition_wss(x_mat, labels),
        silhouette = partition_silhouette(x_mat, labels)
      ) %>%
        bind_cols(cmeans_validity(fit) %>% mutate(m = m))
    } else if (method == "mclust") {
      fit <- fit_mclust_model(x_mat, k = k, seed = seed)
      labels <- fit$classification
      out <- tibble(
        method = method,
        k = k,
        wss = partition_wss(x_mat, labels),
        silhouette = partition_silhouette(x_mat, labels)
      ) %>%
        bind_cols(tibble(BIC = max(fit$bic, na.rm = TRUE), ICL = fit$icl))
    } else {
      stop("Unknown method: ", method)
    }

    out
  })
}

cmeans_fuzziness_diagnostics <- function(x,
                                        k,
                                        m_grid,
                                        seed = 123,
                                        baseline_labels = NULL,
                                        main_threshold = 0.6,
                                        secondary_threshold = 0.7) {
  x_mat <- as.matrix(x)
  m_grid <- sort(unique(m_grid))

  fit_list <- purrr::map(m_grid, function(m) {
    fit <- fit_cmeans_model(x_mat, k = k, m = m, seed = seed)
    list(
      m = m,
      fit = fit,
      labels = max.col(fit$membership)
    )
  })

  purrr::map_dfr(seq_along(fit_list), function(i) {
    current <- fit_list[[i]]
    prev_labels <- if (i == 1) NULL else fit_list[[i - 1]]$labels

    tibble(
      m = current$m,
      k = k,
      wss = partition_wss(x_mat, current$labels),
      silhouette = partition_silhouette(x_mat, current$labels),
      ari_to_previous_m = if (is.null(prev_labels)) {
        NA_real_
      } else {
        mclust::adjustedRandIndex(prev_labels, current$labels)
      },
      ari_to_baseline_kmeans = if (is.null(baseline_labels)) {
        NA_real_
      } else {
        mclust::adjustedRandIndex(as.integer(as.factor(baseline_labels)), current$labels)
      }
    ) %>%
      bind_cols(cmeans_validity(current$fit)) %>%
      bind_cols(
        cmeans_membership_summary(
          current$fit,
          main_threshold = main_threshold,
          secondary_threshold = secondary_threshold
        )
      )
  })
}

plot_diagnostic_curve <- function(diag_df, y_col, title_text, y_lab, method_label = NULL) {
  ggplot(diag_df, aes(x = k, y = .data[[y_col]])) +
    geom_line(linewidth = 0.8, colour = "#1f78b4") +
    geom_point(size = 2, colour = "#1f78b4") +
    scale_x_continuous(breaks = sort(unique(diag_df$k))) +
    labs(
      title = title_text,
      subtitle = method_label,
      x = "Number of clusters",
      y = y_lab
    ) +
    theme_minimal(base_size = 12) +
    theme(
      plot.title = element_text(face = "bold", hjust = 0.5),
      plot.subtitle = element_text(hjust = 0.5)
    )
}

plot_cmeans_m_curve <- function(diag_df, y_col, title_text, y_lab, subtitle_text = NULL) {
  x_limits <- range(diag_df$m, na.rm = TRUE)
  x_breaks <- pretty(x_limits, n = min(8, length(unique(diag_df$m))))

  ggplot(diag_df, aes(x = m, y = .data[[y_col]])) +
    geom_line(linewidth = 0.8, colour = "#e31a1c") +
    geom_point(size = 2, colour = "#e31a1c") +
    scale_x_continuous(
      limits = x_limits,
      breaks = x_breaks
    ) +
    labs(
      title = title_text,
      subtitle = subtitle_text %||% paste0("k = ", unique(diag_df$k)[1]),
      x = "Fuzziness m",
      y = y_lab
    ) +
    theme_minimal(base_size = 12) +
    theme(
      plot.title = element_text(face = "bold", hjust = 0.5),
      plot.subtitle = element_text(hjust = 0.5)
    )
}

plot_cmeans_threshold_curve <- function(diag_df,
                                       title_text,
                                       subtitle_text = NULL,
                                       main_threshold = 0.6,
                                       secondary_threshold = 0.7) {
  threshold_df <- diag_df %>%
    select(m, prop_max_below_main, prop_max_below_secondary) %>%
    pivot_longer(-m, names_to = "metric", values_to = "value") %>%
    mutate(
      metric = recode(
        metric,
        prop_max_below_main = paste0("Max membership < ", format(main_threshold, nsmall = 1)),
        prop_max_below_secondary = paste0("Max membership < ", format(secondary_threshold, nsmall = 1))
      )
    )

  x_limits <- range(diag_df$m, na.rm = TRUE)
  x_breaks <- pretty(x_limits, n = min(8, length(unique(diag_df$m))))

  ggplot(threshold_df, aes(x = m, y = value, colour = metric)) +
    geom_line(linewidth = 0.8) +
    geom_point(size = 2) +
    scale_x_continuous(limits = x_limits, breaks = x_breaks) +
    scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
    scale_colour_manual(
      values = setNames(
        c("#1f78b4", "#33a02c"),
        c(
          paste0("Max membership < ", format(main_threshold, nsmall = 1)),
          paste0("Max membership < ", format(secondary_threshold, nsmall = 1))
        )
      )
    ) +
    labs(
      title = title_text,
      subtitle = subtitle_text %||% paste0("Thresholds: ", format(main_threshold, nsmall = 1), " and ", format(secondary_threshold, nsmall = 1)),
      x = "Fuzziness m",
      y = "Proportion of data zones",
      colour = NULL
    ) +
    theme_minimal(base_size = 12) +
    theme(
      plot.title = element_text(face = "bold", hjust = 0.5),
      plot.subtitle = element_text(hjust = 0.5),
      legend.position = "bottom"
    )
}

count_clusters <- function(x) {
  tibble(cluster = as.character(sort(unique(x)))) %>%
    mutate(n = purrr::map_int(cluster, ~ sum(as.character(x) == .x, na.rm = TRUE)))
}

align_cluster_labels <- function(a, b) {
  a <- as.integer(as.factor(a))
  b <- as.integer(as.factor(b))
  ka <- length(unique(a))
  kb <- length(unique(b))

  if (ka != kb) return(b)
  k <- ka

  perms <- gtools::permutations(n = k, r = k, v = seq_len(k))
  tab <- table(a, b)
  scores <- apply(perms, 1, function(p) {
    sum(tab[cbind(seq_len(k), p)])
  })
  best <- perms[which.max(scores), ]

  mapped <- best[b]
  as.integer(mapped)
}

plot_cluster_map <- function(sf_data, cluster_col, title_text = NULL) {
  x <- as.factor(sf_data[[cluster_col]])
  levs <- levels(x)
  cols <- cluster_palette[levs]

  if (any(is.na(cols))) {
    cols[is.na(cols)] <- grDevices::rainbow(sum(is.na(cols)))
  }

  counts <- count_clusters(x)
  labels <- setNames(
    paste0(counts$cluster, " (n = ", counts$n, ")"),
    counts$cluster
  )

  ggplot(sf_data) +
    geom_sf(aes(fill = as.factor(.data[[cluster_col]])), colour = NA) +
    scale_fill_manual(values = cols, labels = labels, drop = FALSE, name = "Cluster") +
    labs(title = title_text %||% cluster_col) +
    theme_minimal(base_size = 12) +
    theme(
      plot.title = element_text(face = "bold", hjust = 0.5),
      legend.position = "right"
    )
}

plot_difference_map <- function(sf_data, cluster_col_a, cluster_col_b, title_text = NULL, align_labels = TRUE) {
  a <- sf_data[[cluster_col_a]]
  b <- sf_data[[cluster_col_b]]

  if (align_labels) {
    same_k <- length(unique(na.omit(a))) == length(unique(na.omit(b)))
    if (same_k) {
      b <- align_cluster_labels(a, b)
    }
  }

  diff_tbl <- sf_data %>%
    mutate(
      .cluster_a = a,
      .cluster_b = b,
      comparison = case_when(
        is.na(.cluster_a) | is.na(.cluster_b) ~ "Missing",
        .cluster_a == .cluster_b ~ "Same",
        TRUE ~ "Different"
      )
    )

  ggplot(diff_tbl) +
    geom_sf(aes(fill = comparison), colour = NA) +
    scale_fill_manual(
      values = c("Same" = "grey80", "Different" = "#e31a1c", "Missing" = "white"),
      drop = FALSE
    ) +
    labs(
      title = title_text %||% paste(cluster_col_a, "vs", cluster_col_b),
      subtitle = "Difference map after optional label alignment",
      fill = NULL
    ) +
    theme_minimal(base_size = 12) +
    theme(
      plot.title = element_text(face = "bold", hjust = 0.5),
      plot.subtitle = element_text(hjust = 0.5)
    )
}


plot_cmeans_ambiguity_map <- function(sf_data, membership_col, title_text = NULL) {
  ggplot(sf_data) +
    geom_sf(aes(fill = .data[[membership_col]]), colour = NA) +
    scale_fill_viridis_c(
      option = "C",
      direction = 1,
      na.value = "white",
      name = "Max membership"
    ) +
    labs(
      title = title_text %||% membership_col,
      subtitle = "Lower values indicate more ambiguous cluster membership."
    ) +
    theme_minimal(base_size = 12) +
    theme(
      plot.title = element_text(face = "bold", hjust = 0.5),
      plot.subtitle = element_text(hjust = 0.5)
    )
}


plot_cmeans_membership_distribution <- function(membership_df,
                                               membership_col = "max_membership",
                                               title_text = NULL,
                                               main_threshold = 0.6,
                                               secondary_threshold = 0.7) {
  ggplot(membership_df, aes(x = .data[[membership_col]])) +
    geom_histogram(binwidth = 0.02, fill = "#1f78b4", colour = "white", boundary = 0) +
    geom_vline(xintercept = main_threshold, linetype = "dashed", colour = "#e31a1c") +
    geom_vline(xintercept = secondary_threshold, linetype = "dotted", colour = "#33a02c") +
    labs(
      title = title_text %||% "Distribution of maximum memberships",
      subtitle = paste0("Dashed = ", format(main_threshold, nsmall = 1), "; dotted = ", format(secondary_threshold, nsmall = 1), "."),
      x = "Maximum membership",
      y = "Number of data zones"
    ) +
    theme_minimal(base_size = 12) +
    theme(
      plot.title = element_text(face = "bold", hjust = 0.5),
      plot.subtitle = element_text(hjust = 0.5)
    )
}

build_cmeans_membership_export <- function(fit,
                                           datazone,
                                           main_threshold = 0.6) {
  membership_mat <- as.matrix(fit$membership)

  if (ncol(membership_mat) != 2) {
    stop("build_cmeans_membership_export() currently expects k = 2.")
  }

  tibble(
    DataZone = datazone,
    cluster_1_membership = membership_mat[, 1],
    cluster_2_membership = membership_mat[, 2],
    max_membership = apply(membership_mat, 1, max),
    hard_label = max.col(membership_mat)
  ) %>%
    mutate(
      core_borderline_flag = if_else(max_membership < main_threshold, "borderline", "core")
    )
}

plot_domain_boxplots <- function(feature_df,
                                 cluster_col,
                                 id_col = "DataZone",
                                 title_text = NULL,
                                 group_levels = get0(
                                   "group_order",
                                   ifnotfound = c("Rates", "Health", "Access", "Education", "Other"),
                                   inherits = TRUE
                                 )) {
  feature_names <- setdiff(names(feature_df), c(id_col, cluster_col))
  group_map <- group_map_for_features(feature_names, group_levels = group_levels)

  long_df <- feature_df %>%
    pivot_longer(all_of(feature_names), names_to = "Indicator", values_to = "Value") %>%
    left_join(group_map, by = "Indicator") %>%
    mutate(
      cluster = as.factor(.data[[cluster_col]]),
      Group = factor(Group, levels = group_levels),
      Indicator = factor(
        Indicator,
        levels = group_map %>% arrange(Group, Domain, Indicator) %>% pull(Indicator)
      )
    )

  ggplot(long_df, aes(x = Indicator, y = Value, fill = cluster)) +
    geom_boxplot(outlier.size = 0.35, alpha = 0.90) +
    facet_wrap(~ Group, scales = "free_x", ncol = 2) +
    scale_fill_manual(values = cluster_palette, drop = FALSE, name = "Cluster") +
    labs(title = title_text %||% cluster_col, x = NULL, y = "Scaled value") +
    theme_bw(base_size = 11) +
    theme(
      plot.title = element_text(face = "bold", hjust = 0.5),
      strip.text = element_text(face = "bold"),
      axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
      legend.position = "top"
    )
}

selected_similarity_pairs <- function(cluster_cols, method_levels = c("kmeans", "hclust", "cmeans", "mclust")) {
  meta <- tibble(cluster_col = cluster_cols) %>%
    tidyr::extract(
      cluster_col,
      into = c("feature_set", "method", "k"),
      regex = "^(raw|pc)_([A-Za-z]+)_k([0-9]+)$",
      remove = FALSE
    )

  same_feature_diff_method <- meta %>%
    inner_join(meta, by = c("feature_set", "k"), suffix = c("_a", "_b")) %>%
    filter(method_a < method_b)

  same_method_diff_feature <- meta %>%
    inner_join(meta, by = c("method", "k"), suffix = c("_a", "_b")) %>%
    filter(feature_set_a == "raw", feature_set_b == "pc")

  bind_rows(
    same_feature_diff_method %>%
      transmute(clustering_a = cluster_col_a, clustering_b = cluster_col_b),
    same_method_diff_feature %>%
      transmute(clustering_a = cluster_col_a, clustering_b = cluster_col_b)
  ) %>%
    distinct()
}

compute_similarity_table <- function(assignments, pairs_df) {
  purrr::pmap_dfr(pairs_df, function(clustering_a, clustering_b) {
    tibble(
      clustering_a = clustering_a,
      clustering_b = clustering_b,
      ARI = mclust::adjustedRandIndex(assignments[[clustering_a]], assignments[[clustering_b]]),
      NMI = aricode::NMI(assignments[[clustering_a]], assignments[[clustering_b]]),
      VI = salso::VI(assignments[[clustering_a]], assignments[[clustering_b]])
    )
  })
}

pretty_cluster_label <- function(x) {
  x %>%
    stringr::str_replace("^raw_", "raw\n") %>%
    stringr::str_replace("^pc_", "pc\n") %>%
    stringr::str_replace("_k", "\nk")
}

plot_similarity_heatmap <- function(similarity_df, metric = c("ARI", "NMI", "VI"), title_text = NULL) {
  metric <- match.arg(metric)

  nodes <- sort(unique(c(similarity_df$clustering_a, similarity_df$clustering_b)))
  mat <- matrix(NA_real_, nrow = length(nodes), ncol = length(nodes), dimnames = list(nodes, nodes))
  diag(mat) <- if (metric == "VI") 0 else 1

  for (i in seq_len(nrow(similarity_df))) {
    a <- similarity_df$clustering_a[i]
    b <- similarity_df$clustering_b[i]
    val <- similarity_df[[metric]][i]
    mat[a, b] <- val
    mat[b, a] <- val
  }

  heat_df <- as.data.frame(as.table(mat), stringsAsFactors = FALSE) %>%
    as_tibble() %>%
    rename(clustering_row = Var1, clustering_col = Var2, value = Freq) %>%
    mutate(
      clustering_row = factor(clustering_row, levels = rev(nodes), labels = pretty_cluster_label(rev(nodes))),
      clustering_col = factor(clustering_col, levels = nodes, labels = pretty_cluster_label(nodes)),
      label = if_else(is.na(value), "", sprintf("%.3f", value))
    )

  fill_limits <- if (metric == "VI") c(0, max(heat_df$value, na.rm = TRUE)) else c(0, 1)
  fill_midpoint <- if (metric == "VI") mean(fill_limits) else 0.5

  ggplot(heat_df, aes(x = clustering_col, y = clustering_row, fill = value)) +
    geom_tile(colour = "white", linewidth = 0.4) +
    geom_text(aes(label = label), size = 3.4) +
    scale_fill_gradient2(
      low = "#4575b4",
      mid = "white",
      high = "#d73027",
      midpoint = fill_midpoint,
      limits = fill_limits,
      na.value = "grey92"
    ) +
    coord_equal() +
    labs(
      title = title_text %||% metric,
      x = NULL,
      y = NULL,
      fill = metric
    ) +
    theme_minimal(base_size = 11) +
    theme(
      panel.grid = element_blank(),
      axis.text.x = element_text(angle = 0, hjust = 0.5, vjust = 0.5, size = 10),
      axis.text.y = element_text(size = 10),
      plot.title = element_text(face = "bold", hjust = 0.5),
      legend.position = "right"
    )
}

make_metric_matrix <- function(similarity_df, metric = c("ARI", "NMI", "VI")) {
  metric <- match.arg(metric)

  nodes <- sort(unique(c(similarity_df$clustering_a, similarity_df$clustering_b)))
  mat <- matrix(NA_real_, nrow = length(nodes), ncol = length(nodes), dimnames = list(nodes, nodes))
  diag(mat) <- if (metric == "VI") 0 else 1

  for (i in seq_len(nrow(similarity_df))) {
    a <- similarity_df$clustering_a[i]
    b <- similarity_df$clustering_b[i]
    val <- similarity_df[[metric]][i]
    mat[a, b] <- val
    mat[b, a] <- val
  }

  as_tibble(mat, rownames = "clustering_result")
}

pick_cluster_col <- function(feature_set, method, k) {
  paste0(feature_set, "_", method, "_k", k)
}
```

```{r extra-package-check}
if (!requireNamespace("gtools", quietly = TRUE)) {
  stop("Install package 'gtools' before knitting.")
}
```

# Data preprocessing

## Read and filter the data

```{r read-data}
csv_path <- file.path(params$data_dir, params$csv_file)
shp_path <- file.path(params$data_dir, params$shp_file)

assert_file_exists(csv_path)
assert_file_exists(shp_path)

simd_tbl <- readr::read_csv(csv_path, show_col_types = FALSE)
simd_shp <- sf::st_read(shp_path, quiet = TRUE)

simd_joined <- simd_shp %>%
  left_join(simd_tbl, by = c("DataZone" = "Data_Zone")) %>%
  left_join(joint_board_lookup, by = "Council_area")

overall_rank_source_col <- first_existing_col(
  simd_joined,
  c(
    "SIMD2020v2_Rank",
    "SIMD2020_Rank",
    "SIMD_Rank",
    "Overall_SIMD_Rank",
    "overall_simd_rank",
    "simd_rank"
  )
)

if (is.null(overall_rank_source_col)) {
  board_meta <- simd_joined %>%
    filter(Joint_Board_area == params$target_board) %>%
    st_drop_geometry() %>%
    transmute(DataZone, overall_simd_rank = NA_real_)
} else {
  board_meta <- simd_joined %>%
    filter(Joint_Board_area == params$target_board) %>%
    st_drop_geometry() %>%
    transmute(DataZone, overall_simd_rank = .data[[overall_rank_source_col]])
}

board_sf <- simd_joined %>%
  filter(Joint_Board_area == params$target_board) %>%
  select(DataZone, geometry, any_of(indicator_cols))

board_raw <- board_sf %>%
  st_drop_geometry() %>%
  clean_percent_numeric(id_col = "DataZone")
```

## Missingness

```{r missingness}
missing_summary <- make_missing_summary(board_raw, id_col = "DataZone")
missing_summary_nonzero <- missing_summary %>%
  filter(Missing_Count > 0)

if (nrow(missing_summary_nonzero) == 0) {
  cat("No indicators have missing values in this board after reading the source file.")
} else {
  kable(
    missing_summary_nonzero,
    digits = 2,
    caption = paste0("Indicators with missing values in ", params$target_board)
  ) %>%
    kable_styling(full_width = FALSE)
}
```

```{r choose-indicators}
kept_indicators <- missing_summary %>%
  filter(Percent_Missing <= params$missing_threshold) %>%
  pull(Indicator)

dropped_indicators <- setdiff(indicator_cols, kept_indicators)

board_complete <- board_raw %>%
  select(DataZone, all_of(kept_indicators)) %>%
  tidyr::drop_na()

board_sf_complete <- board_sf %>%
  semi_join(board_complete, by = "DataZone")

cat("Indicators retained after missingness filtering:", length(kept_indicators), "

")
cat("Indicators dropped:", ifelse(length(dropped_indicators) == 0, "None", paste(dropped_indicators, collapse = ", ")))
```

## Statistics

```{r statistics}
analysis_stats_tbl <- format_stats_table(board_complete, id_col = "DataZone")

kable(
  analysis_stats_tbl,
  caption = paste0("Descriptive statistics for ", params$target_board, " analysis data")
) %>%
  kable_styling(full_width = FALSE)
```

## Distribution

```{r distributions, echo=FALSE, fig.width=11, fig.height=9}
plot_domain_distributions(
  board_complete,
  id_col = "DataZone",
  title_text = paste0("Distribution of ", params$target_board, " indicators"),
  by = "domain"
)
```

## Correlation

```{r correlation-all, echo=FALSE, fig.width=11, fig.height=9}
plot_correlation_heatmap(
  board_complete,
  id_col = "DataZone",
  title_text = paste0("Correlation heatmap of ", params$target_board, " indicators")
)
```

## PCA

The two clustering feature sets keep the rate variables for Income, Employment, Crime, and Housing in both designs.

For the remaining domains:

- the **raw-indicator-based** set keeps all raw indicators with missingness below the threshold;
- the **PC-based** set replaces the remaining domains with domain-wise PCs whose **individual variance explained is greater than 10%**.

```{r build-feature-sets}
available_domains <- indicator_domains %>%
  filter(Indicator %in% kept_indicators) %>%
  mutate(Domain = factor(Domain, levels = domain_order))

remaining_domains <- available_domains %>%
  filter(!Domain %in% core_domains) %>%
  distinct(Domain) %>%
  pull(Domain) %>%
  as.character()

available_core_rate_vars <- intersect(core_rate_vars, names(board_complete))

raw_remaining_vars <- available_domains %>%
  filter(!Domain %in% core_domains) %>%
  pull(Indicator)

raw_feature_tbl <- board_complete %>%
  select(DataZone, all_of(c(available_core_rate_vars, raw_remaining_vars))) %>%
  scale_df(id_col = "DataZone")

pca_input_tbl <- board_complete %>%
  select(DataZone, all_of(unique(c(available_core_rate_vars, raw_remaining_vars)))) %>%
  scale_df(id_col = "DataZone")

pca_results <- run_domain_pca(
  df = pca_input_tbl %>% select(-DataZone),
  domain_map = available_domains,
  domains = remaining_domains,
  variance_cutoff = 0.10
)

variance_tbl <- bind_rows(purrr::map(pca_results, "variance"))
loadings_tbl <- bind_rows(purrr::map(pca_results, "loadings"), .id = "Domain_Source")
retained_pc_scores <- purrr::map(pca_results, "retained_scores") %>% purrr::compact()

pc_feature_tbl <- board_complete %>%
  select(DataZone, all_of(available_core_rate_vars)) %>%
  scale_df(id_col = "DataZone")

if (length(retained_pc_scores) > 0) {
  pc_feature_tbl <- bind_cols(
    pc_feature_tbl %>% select(DataZone),
    pc_feature_tbl %>% select(-DataZone),
    bind_cols(retained_pc_scores)
  )
}

raw_feature_names <- setdiff(names(raw_feature_tbl), "DataZone")
pc_feature_names <- setdiff(names(pc_feature_tbl), "DataZone")
```

\newpage

```{r feature-dictionaries, results='asis'}
feature_representation_tbl <- feature_representation_summary(raw_feature_names, pc_feature_names)

kable(
  feature_representation_tbl,
  caption = "Domain-wise comparison of variables retained in the raw and PC-based feature sets",
  booktabs = TRUE,
  longtable = TRUE
) %>%
  kable_styling(full_width = FALSE, latex_options = c("repeat_header"))
```

```{r pca-variance, fig.width=12, fig.height=8}
if (nrow(variance_tbl) > 0) {
  var_plots <- variance_tbl %>%
    split(.$Domain) %>%
    purrr::imap(~ plot_domain_variance(.x, paste(.y, "domain PCA")))

  wrap_plots(var_plots) +
    plot_annotation(
      title = "Variance explained by domain-wise principal components",
      subtitle = "PCs retained when individual variance explained is greater than 10%"
    )
}
```

```{r pca-loadings, results='asis', echo=FALSE}
if (nrow(loadings_tbl) > 0) {
  split_loadings <- split(loadings_tbl, loadings_tbl$Domain_Source)

  purrr::iwalk(split_loadings, function(tbl, nm) {
    tbl_out <- tbl %>%
      select(Indicator, starts_with(paste0(nm, "_")))
    cat("\n\n")
    print(
      kable(
        tbl_out,
        digits = 3,
        caption = paste0("PCA loadings for retained ", nm, " components")
      ) %>%
        kable_styling(full_width = FALSE)
    )
    cat("\n\n")
  })
}
```

```{r raw-feature-distributions, echo=FALSE, fig.width=10, fig.height=9}
plot_domain_distributions(
  raw_feature_tbl,
  id_col = "DataZone",
  title_text = "Distribution of raw-indicator-based clustering variables",
  geom_type = "histogram",
  ncol = 4,
  bins = 20,
  by = "domain"
)
```

```{r pc-feature-distributions, echo=FALSE, fig.width=10, fig.height=7}
plot_domain_distributions(
  pc_feature_tbl,
  id_col = "DataZone",
  title_text = "Distribution of PC-based clustering variables",
  geom_type = "histogram",
  ncol = 4,
  bins = 20,
  by = "domain"
)
```

```{r feature-set-correlations, echo=FALSE, fig.width=11, fig.height=10}
p_raw_corr <- plot_correlation_heatmap(
  raw_feature_tbl,
  id_col = "DataZone",
  title_text = "Correlation heatmap: raw-indicator-based feature set"
)

p_pc_corr <- plot_correlation_heatmap(
  pc_feature_tbl,
  id_col = "DataZone",
  title_text = "Correlation heatmap: PC-based feature set"
)

wrap_plots(p_raw_corr, p_pc_corr, ncol = 1)
```


# Clustering

```{r clustering-inputs, include=FALSE}
raw_x <- as.matrix(raw_feature_tbl %>% select(-DataZone))
pc_x <- as.matrix(pc_feature_tbl %>% select(-DataZone))

selected_k_raw <- params$chosen_k_raw
selected_k_pc <- params$chosen_k_pc
reference_m_raw <- params$cmeans_ref_m_raw
reference_m_pc <- params$cmeans_ref_m_pc
```

## K-means

### Raw-based clustering

```{r kmeans-raw-diagnostics, fig.width=13, fig.height=6}
diag_kmeans_raw <- method_diagnostics(raw_x, params$k_grid, method = "kmeans", seed = params$seed) %>%
  mutate(feature_set = "raw")

p_kmeans_raw_wss <- plot_diagnostic_curve(
  diag_kmeans_raw,
  y_col = "wss",
  title_text = "K-means elbow plot",
  y_lab = "Within-cluster sum of squares",
  method_label = "Raw-indicator-based feature set"
)

p_kmeans_raw_sil <- plot_diagnostic_curve(
  diag_kmeans_raw,
  y_col = "silhouette",
  title_text = "K-means silhouette plot",
  y_lab = "Average silhouette width",
  method_label = "Raw-indicator-based feature set"
)

wrap_plots(p_kmeans_raw_wss, p_kmeans_raw_sil, ncol = 2)
```

```{r kmeans-raw-fit}
raw_kmeans_col <- pick_cluster_col("raw", "kmeans", selected_k_raw)

cluster_assignments <- tibble(DataZone = board_complete$DataZone)
cluster_assignments[[raw_kmeans_col]] <- fit_kmeans_labels(raw_x, k = selected_k_raw, seed = params$seed)

raw_kmeans_features <- raw_feature_tbl %>%
  left_join(cluster_assignments %>% select(DataZone, all_of(raw_kmeans_col)), by = "DataZone")

kable(
  count_clusters(raw_kmeans_features[[raw_kmeans_col]]),
  caption = paste0("Cluster sizes for ", caption_safe(raw_kmeans_col))
) %>%
  kable_styling(full_width = FALSE)
```

```{r kmeans-raw-boxplot, fig.width=13, fig.height=9}
plot_domain_boxplots(
  raw_kmeans_features,
  cluster_col = raw_kmeans_col,
  title_text = paste0("K-means, raw feature set: ", raw_kmeans_col)
)
```

### PC-based clustering

```{r kmeans-pc-diagnostics, fig.width=13, fig.height=6}
diag_kmeans_pc <- method_diagnostics(pc_x, params$k_grid, method = "kmeans", seed = params$seed) %>%
  mutate(feature_set = "pc")

p_kmeans_pc_wss <- plot_diagnostic_curve(
  diag_kmeans_pc,
  y_col = "wss",
  title_text = "K-means elbow plot",
  y_lab = "Within-cluster sum of squares",
  method_label = "PC-based feature set"
)

p_kmeans_pc_sil <- plot_diagnostic_curve(
  diag_kmeans_pc,
  y_col = "silhouette",
  title_text = "K-means silhouette plot",
  y_lab = "Average silhouette width",
  method_label = "PC-based feature set"
)

wrap_plots(p_kmeans_pc_wss, p_kmeans_pc_sil, ncol = 2)
```

```{r kmeans-pc-fit}
pc_kmeans_col <- pick_cluster_col("pc", "kmeans", selected_k_pc)
cluster_assignments[[pc_kmeans_col]] <- fit_kmeans_labels(pc_x, k = selected_k_pc, seed = params$seed)

pc_kmeans_features <- pc_feature_tbl %>%
  left_join(cluster_assignments %>% select(DataZone, all_of(pc_kmeans_col)), by = "DataZone")

kable(
  count_clusters(pc_kmeans_features[[pc_kmeans_col]]),
  caption = paste0("Cluster sizes for ", caption_safe(pc_kmeans_col))
) %>%
  kable_styling(full_width = FALSE)
```

```{r kmeans-pc-boxplot, fig.width=13, fig.height=9}
plot_domain_boxplots(
  pc_kmeans_features,
  cluster_col = pc_kmeans_col,
  title_text = paste0("K-means, PC feature set: ", pc_kmeans_col)
)
```

### Comparison

```{r kmeans-comparison}
kmeans_comparison_tbl <- tibble(
  comparison = c("raw", "pc"),
  selected_k = c(selected_k_raw, selected_k_pc),
  best_silhouette_k = c(
    best_metric_k(diag_kmeans_raw, "silhouette", "k"),
    best_metric_k(diag_kmeans_pc, "silhouette", "k")
  ),
  best_silhouette = c(
    best_metric_value(diag_kmeans_raw, "silhouette"),
    best_metric_value(diag_kmeans_pc, "silhouette")
  )
)

kable(
  kmeans_comparison_tbl,
  digits = 3,
  caption = "K-means comparison between raw and PC feature sets"
) %>%
  kable_styling(full_width = FALSE)
```

## Hierarchical clustering

### Raw-based clustering

```{r hclust-raw-diagnostics, fig.width=13, fig.height=6}
diag_hclust_raw <- method_diagnostics(raw_x, params$k_grid, method = "hierarchical", seed = params$seed) %>%
  mutate(feature_set = "raw")

p_hclust_raw_wss <- plot_diagnostic_curve(
  diag_hclust_raw,
  y_col = "wss",
  title_text = "Hierarchical elbow plot",
  y_lab = "Within-cluster sum of squares",
  method_label = "Raw-indicator-based feature set"
)

p_hclust_raw_sil <- plot_diagnostic_curve(
  diag_hclust_raw,
  y_col = "silhouette",
  title_text = "Hierarchical silhouette plot",
  y_lab = "Average silhouette width",
  method_label = "Raw-indicator-based feature set"
)

wrap_plots(p_hclust_raw_wss, p_hclust_raw_sil, ncol = 2)
```

```{r hclust-raw-fit}
raw_hclust_col <- pick_cluster_col("raw", "hclust", selected_k_raw)
cluster_assignments[[raw_hclust_col]] <- fit_hclust_labels(raw_x, k = selected_k_raw)

raw_hclust_features <- raw_feature_tbl %>%
  left_join(cluster_assignments %>% select(DataZone, all_of(raw_hclust_col)), by = "DataZone")

kable(
  count_clusters(raw_hclust_features[[raw_hclust_col]]),
  caption = paste0("Cluster sizes for ", caption_safe(raw_hclust_col))
) %>%
  kable_styling(full_width = FALSE)
```

```{r hclust-raw-boxplot, fig.width=13, fig.height=9}
plot_domain_boxplots(
  raw_hclust_features,
  cluster_col = raw_hclust_col,
  title_text = paste0("Hierarchical clustering, raw feature set: ", raw_hclust_col)
)
```

### PC-based clustering

```{r hclust-pc-diagnostics, fig.width=13, fig.height=6}
diag_hclust_pc <- method_diagnostics(pc_x, params$k_grid, method = "hierarchical", seed = params$seed) %>%
  mutate(feature_set = "pc")

p_hclust_pc_wss <- plot_diagnostic_curve(
  diag_hclust_pc,
  y_col = "wss",
  title_text = "Hierarchical elbow plot",
  y_lab = "Within-cluster sum of squares",
  method_label = "PC-based feature set"
)

p_hclust_pc_sil <- plot_diagnostic_curve(
  diag_hclust_pc,
  y_col = "silhouette",
  title_text = "Hierarchical silhouette plot",
  y_lab = "Average silhouette width",
  method_label = "PC-based feature set"
)

wrap_plots(p_hclust_pc_wss, p_hclust_pc_sil, ncol = 2)
```

```{r hclust-pc-fit}
pc_hclust_col <- pick_cluster_col("pc", "hclust", selected_k_pc)
cluster_assignments[[pc_hclust_col]] <- fit_hclust_labels(pc_x, k = selected_k_pc)

pc_hclust_features <- pc_feature_tbl %>%
  left_join(cluster_assignments %>% select(DataZone, all_of(pc_hclust_col)), by = "DataZone")

kable(
  count_clusters(pc_hclust_features[[pc_hclust_col]]),
  caption = paste0("Cluster sizes for ", caption_safe(pc_hclust_col))
) %>%
  kable_styling(full_width = FALSE)
```

```{r hclust-pc-boxplot, fig.width=13, fig.height=9}
plot_domain_boxplots(
  pc_hclust_features,
  cluster_col = pc_hclust_col,
  title_text = paste0("Hierarchical clustering, PC feature set: ", pc_hclust_col)
)
```

### Comparison

```{r hclust-comparison}
hclust_comparison_tbl <- tibble(
  comparison = c("raw", "pc"),
  selected_k = c(selected_k_raw, selected_k_pc),
  best_silhouette_k = c(
    best_metric_k(diag_hclust_raw, "silhouette", "k"),
    best_metric_k(diag_hclust_pc, "silhouette", "k")
  ),
  best_silhouette = c(
    best_metric_value(diag_hclust_raw, "silhouette"),
    best_metric_value(diag_hclust_pc, "silhouette")
  )
)

kable(
  hclust_comparison_tbl,
  digits = 3,
  caption = "Hierarchical clustering comparison between raw and PC feature sets"
) %>%
  kable_styling(full_width = FALSE)
```

## C-means (sensitivity / uncertainty layer)

C-means is retained here as a **sensitivity / uncertainty layer** around the headline two-cluster K-means result. It is used to show which data zones are **core members** of each profile and which are **boundary cases**, not to reopen the choice of `k` and not to claim a sharp optimum in `m`.

Accordingly, `k` is held fixed at the chosen K-means value for each feature set, and the fuzziness exponent `m` is varied over a practical range (`params$cmeans_m_grid`, default `1.1` to `2.5`). The **reference/reporting value `m = 2`** is used for the headline C-means summaries, not as a strongly selected optimum. The main C-means diagnostics are therefore:

- mean maximum membership,
- proportion with maximum membership below `r params$cmeans_main_ambiguity_threshold` (main ambiguity threshold),
- proportion with maximum membership below `r params$cmeans_secondary_ambiguity_threshold` (stricter secondary threshold),
- ARI between adjacent `m` values,
- ARI between hardened C-means labels and the baseline K-means labels.

Silhouette and WSS are kept only as **secondary compactness summaries** of hardened assignments.

### Raw-based clustering

```{r cmeans-raw-sensitivity}
if (isTRUE(params$use_cmeans)) {
  cmeans_m_diag_raw <- cmeans_fuzziness_diagnostics(
    raw_x,
    k = selected_k_raw,
    m_grid = params$cmeans_m_grid,
    seed = params$seed,
    baseline_labels = cluster_assignments[[raw_kmeans_col]],
    main_threshold = params$cmeans_main_ambiguity_threshold,
    secondary_threshold = params$cmeans_secondary_ambiguity_threshold
  )

  p_cmeans_raw_meanmax <- plot_cmeans_m_curve(
    cmeans_m_diag_raw,
    y_col = "mean_max_membership",
    title_text = "Mean max membership",
    y_lab = "Mean maximum membership",
    subtitle_text = "Raw feature set"
  )

  p_cmeans_raw_prop <- plot_cmeans_threshold_curve(
    cmeans_m_diag_raw,
    title_text = "Ambiguity shares",
    subtitle_text = "Raw feature set",
    main_threshold = params$cmeans_main_ambiguity_threshold,
    secondary_threshold = params$cmeans_secondary_ambiguity_threshold
  )

  p_cmeans_raw_ari_prev <- plot_cmeans_m_curve(
    cmeans_m_diag_raw,
    y_col = "ari_to_previous_m",
    title_text = "ARI to previous m",
    y_lab = "ARI to previous m",
    subtitle_text = "Raw feature set"
  )

  p_cmeans_raw_ari_base <- plot_cmeans_m_curve(
    cmeans_m_diag_raw,
    y_col = "ari_to_baseline_kmeans",
    title_text = "ARI to K-means",
    y_lab = "ARI to baseline K-means",
    subtitle_text = "Raw feature set"
  )

  ((p_cmeans_raw_meanmax | p_cmeans_raw_prop) / (p_cmeans_raw_ari_prev | p_cmeans_raw_ari_base)) +
    patchwork::plot_annotation(title = paste0("Raw C-means sensitivity (k = ", selected_k_raw, ")"))
}
```

```{r cmeans-raw-fit}
if (isTRUE(params$use_cmeans)) {
  raw_cmeans_col <- pick_cluster_col("raw", "cmeans", selected_k_raw)
  raw_cmeans_membership_col <- paste0(raw_cmeans_col, "_max_membership")

  raw_cmeans_fit <- fit_cmeans_model(
    raw_x,
    k = selected_k_raw,
    m = reference_m_raw,
    seed = params$seed
  )

  raw_cmeans_membership_export <- build_cmeans_membership_export(
    raw_cmeans_fit,
    datazone = raw_feature_tbl$DataZone,
    main_threshold = params$cmeans_main_ambiguity_threshold
  )

  raw_cmeans_membership_tbl <- raw_cmeans_membership_export %>%
    transmute(
      DataZone,
      !!raw_cmeans_col := hard_label,
      !!raw_cmeans_membership_col := max_membership
    )

  cluster_assignments[[raw_cmeans_col]] <- raw_cmeans_membership_tbl[[raw_cmeans_col]]

  raw_cmeans_features <- raw_feature_tbl %>%
    left_join(
      raw_cmeans_membership_tbl %>% select(DataZone, all_of(raw_cmeans_col)),
      by = "DataZone"
    )

  raw_cmeans_reference_summary <- cmeans_m_diag_raw %>%
    slice(which.min(abs(m - reference_m_raw))) %>%
    transmute(
      `k` = k,
      `Reference/reporting value m` = m,
      `Mean max membership` = mean_max_membership,
      `Prop. max < 0.6` = prop_max_below_main,
      `Prop. max < 0.7` = prop_max_below_secondary,
      `ARI to previous m` = ari_to_previous_m,
      `ARI to K-means` = ari_to_baseline_kmeans,
      `Silhouette (hard labels)` = silhouette
    )

  kable(
    raw_cmeans_reference_summary,
    digits = 3,
    caption = "Reference C-means summary for the raw feature set"
  ) %>%
    kable_styling(full_width = FALSE)

  kable(
    count_clusters(raw_cmeans_features[[raw_cmeans_col]]),
    caption = paste0("Cluster sizes for ", caption_safe(raw_cmeans_col))
  ) %>%
    kable_styling(full_width = FALSE)
}
```

```{r cmeans-raw-boxplot, fig.width=13, fig.height=9}
if (isTRUE(params$use_cmeans)) {
  plot_domain_boxplots(
    raw_cmeans_features,
    cluster_col = raw_cmeans_col,
    title_text = paste0("C-means reference solution, raw feature set: ", raw_cmeans_col, " (reference/reporting value m = ", reference_m_raw, ")")
  )
}
```

```{r cmeans-raw-ambiguity-map, fig.width=8, fig.height=6}
if (isTRUE(params$use_cmeans)) {
  raw_cmeans_map_sf <- board_sf_complete %>%
    select(DataZone, geometry) %>%
    left_join(
      raw_cmeans_membership_tbl %>% select(DataZone, all_of(raw_cmeans_membership_col)),
      by = "DataZone"
    )

  plot_cmeans_ambiguity_map(
    raw_cmeans_map_sf,
    membership_col = raw_cmeans_membership_col,
    title_text = paste0("Most ambiguous areas in the raw C-means solution at the reference/reporting value m = ", reference_m_raw, "")
  )
}
```

### PC-based clustering

```{r cmeans-pc-sensitivity}
if (isTRUE(params$use_cmeans)) {
  cmeans_m_diag_pc <- cmeans_fuzziness_diagnostics(
    pc_x,
    k = selected_k_pc,
    m_grid = params$cmeans_m_grid,
    seed = params$seed,
    baseline_labels = cluster_assignments[[pc_kmeans_col]],
    main_threshold = params$cmeans_main_ambiguity_threshold,
    secondary_threshold = params$cmeans_secondary_ambiguity_threshold
  )

  p_cmeans_pc_meanmax <- plot_cmeans_m_curve(
    cmeans_m_diag_pc,
    y_col = "mean_max_membership",
    title_text = "Mean max membership",
    y_lab = "Mean maximum membership",
    subtitle_text = "PC feature set"
  )

  p_cmeans_pc_prop <- plot_cmeans_threshold_curve(
    cmeans_m_diag_pc,
    title_text = "Ambiguity shares",
    subtitle_text = "PC feature set",
    main_threshold = params$cmeans_main_ambiguity_threshold,
    secondary_threshold = params$cmeans_secondary_ambiguity_threshold
  )

  p_cmeans_pc_ari_prev <- plot_cmeans_m_curve(
    cmeans_m_diag_pc,
    y_col = "ari_to_previous_m",
    title_text = "ARI to previous m",
    y_lab = "ARI to previous m",
    subtitle_text = "PC feature set"
  )

  p_cmeans_pc_ari_base <- plot_cmeans_m_curve(
    cmeans_m_diag_pc,
    y_col = "ari_to_baseline_kmeans",
    title_text = "ARI to K-means",
    y_lab = "ARI to baseline K-means",
    subtitle_text = "PC feature set"
  )

  ((p_cmeans_pc_meanmax | p_cmeans_pc_prop) / (p_cmeans_pc_ari_prev | p_cmeans_pc_ari_base)) +
    patchwork::plot_annotation(title = paste0("PC C-means sensitivity (k = ", selected_k_pc, ")"))
}
```

```{r cmeans-pc-fit}
if (isTRUE(params$use_cmeans)) {
  pc_cmeans_col <- pick_cluster_col("pc", "cmeans", selected_k_pc)
  pc_cmeans_membership_col <- paste0(pc_cmeans_col, "_max_membership")

  pc_cmeans_fit <- fit_cmeans_model(
    pc_x,
    k = selected_k_pc,
    m = reference_m_pc,
    seed = params$seed
  )

  pc_cmeans_membership_export <- build_cmeans_membership_export(
    pc_cmeans_fit,
    datazone = pc_feature_tbl$DataZone,
    main_threshold = params$cmeans_main_ambiguity_threshold
  )

  pc_cmeans_membership_tbl <- pc_cmeans_membership_export %>%
    transmute(
      DataZone,
      !!pc_cmeans_col := hard_label,
      !!pc_cmeans_membership_col := max_membership
    )

  cluster_assignments[[pc_cmeans_col]] <- pc_cmeans_membership_tbl[[pc_cmeans_col]]

  pc_cmeans_features <- pc_feature_tbl %>%
    left_join(
      pc_cmeans_membership_tbl %>% select(DataZone, all_of(pc_cmeans_col)),
      by = "DataZone"
    )

  pc_cmeans_reference_summary <- cmeans_m_diag_pc %>%
    slice(which.min(abs(m - reference_m_pc))) %>%
    transmute(
      `k` = k,
      `Reference/reporting value m` = m,
      `Mean max membership` = mean_max_membership,
      `Prop. max < 0.6` = prop_max_below_main,
      `Prop. max < 0.7` = prop_max_below_secondary,
      `ARI to previous m` = ari_to_previous_m,
      `ARI to K-means` = ari_to_baseline_kmeans,
      `Silhouette (hard labels)` = silhouette
    )

  kable(
    pc_cmeans_reference_summary,
    digits = 3,
    caption = "Reference C-means summary for the PC feature set"
  ) %>%
    kable_styling(full_width = FALSE)

  kable(
    count_clusters(pc_cmeans_features[[pc_cmeans_col]]),
    caption = paste0("Cluster sizes for ", caption_safe(pc_cmeans_col))
  ) %>%
    kable_styling(full_width = FALSE)
}
```

```{r cmeans-pc-boxplot, fig.width=13, fig.height=9}
if (isTRUE(params$use_cmeans)) {
  plot_domain_boxplots(
    pc_cmeans_features,
    cluster_col = pc_cmeans_col,
    title_text = paste0("C-means reference solution, PC feature set: ", pc_cmeans_col, " (reference/reporting value m = ", reference_m_pc, ")")
  )
}
```

```{r cmeans-pc-ambiguity-map, fig.width=8, fig.height=6}
if (isTRUE(params$use_cmeans)) {
  pc_cmeans_map_sf <- board_sf_complete %>%
    select(DataZone, geometry) %>%
    left_join(
      pc_cmeans_membership_tbl %>% select(DataZone, all_of(pc_cmeans_membership_col)),
      by = "DataZone"
    )

  plot_cmeans_ambiguity_map(
    pc_cmeans_map_sf,
    membership_col = pc_cmeans_membership_col,
    title_text = paste0("Most ambiguous areas in the PC C-means solution at the reference/reporting value m = ", reference_m_pc, "")
  )
}
```

### Reference-solution comparison

For this comparison, `k` is fixed at 2 and the **reference/reporting value `m = 2`** is used for both feature sets.

```{r cmeans-comparison}
if (isTRUE(params$use_cmeans)) {
  cmeans_comparison_long_tbl <- bind_rows(
    cmeans_m_diag_raw %>%
      slice(which.min(abs(m - reference_m_raw))) %>%
      transmute(
        feature_set = "Raw",
        fixed_k = k,
        reference_reporting_value_m = m,
        `Mean max membership` = mean_max_membership,
        `Prop. max < 0.6` = prop_max_below_main,
        `Prop. max < 0.7` = prop_max_below_secondary,
        `ARI to previous m` = ari_to_previous_m,
        `ARI to K-means` = ari_to_baseline_kmeans,
        `Silhouette (hard labels)` = silhouette
      ),
    cmeans_m_diag_pc %>%
      slice(which.min(abs(m - reference_m_pc))) %>%
      transmute(
        feature_set = "PC",
        fixed_k = k,
        reference_reporting_value_m = m,
        `Mean max membership` = mean_max_membership,
        `Prop. max < 0.6` = prop_max_below_main,
        `Prop. max < 0.7` = prop_max_below_secondary,
        `ARI to previous m` = ari_to_previous_m,
        `ARI to K-means` = ari_to_baseline_kmeans,
        `Silhouette (hard labels)` = silhouette
      )
  ) %>%
    tidyr::pivot_longer(
      cols = -c(feature_set, fixed_k, reference_reporting_value_m),
      names_to = "summary",
      values_to = "value"
    ) %>%
    mutate(
      summary = factor(
        summary,
        levels = c(
          "Mean max membership",
          "Prop. max < 0.6",
          "Prop. max < 0.7",
          "ARI to previous m",
          "ARI to K-means",
          "Silhouette (hard labels)"
        )
      )
    ) %>%
    arrange(feature_set, summary) %>%
    select(`Feature set` = feature_set, `k` = fixed_k, `Reference/reporting value m` = reference_reporting_value_m, Summary = summary, Value = value)

  kable(
    cmeans_comparison_long_tbl,
    digits = 3,
    caption = "Reference C-means summary by feature set"
  ) %>%
    kable_styling(full_width = FALSE)
}

```

At the reference/reporting value `m = 2`, the PC-based two-cluster solution is sharper, while the raw solution contains more boundary cases.

```{r cmeans-reference-membership-distributions, fig.width=12, fig.height=5.5}
if (isTRUE(params$use_cmeans)) {
  p_raw_membership_dist <- plot_cmeans_membership_distribution(
    raw_cmeans_membership_export,
    title_text = paste0("Raw solution at the reference/reporting value m = ", reference_m_raw, ""),
    main_threshold = params$cmeans_main_ambiguity_threshold,
    secondary_threshold = params$cmeans_secondary_ambiguity_threshold
  )

  p_pc_membership_dist <- plot_cmeans_membership_distribution(
    pc_cmeans_membership_export,
    title_text = paste0("PC solution at the reference/reporting value m = ", reference_m_pc, ""),
    main_threshold = params$cmeans_main_ambiguity_threshold,
    secondary_threshold = params$cmeans_secondary_ambiguity_threshold
  )

  p_raw_membership_dist | p_pc_membership_dist
}
```

## Finite mixture models (secondary model-based comparator)

Finite mixture models are kept here as a **secondary model-based comparator**, not as co-equal evidence for the same two-cluster story. In these data, the BIC-preferred component count can differ materially from `k = 2`, so the two-component FMM should be read as a **constrained comparator for consistency**, not as independent support for the headline two-cluster conclusion.

BIC and ICL are therefore the main FMM diagnostics. Silhouette is shown only as a **secondary compactness summary** of the hardened assignments.

### Raw-based clustering

```{r mclust-raw-diagnostics}
if (isTRUE(params$use_mclust)) {
  diag_mclust_raw <- method_diagnostics(raw_x, params$k_grid, method = "mclust", seed = params$seed) %>%
    mutate(feature_set = "raw")

  p_mclust_raw_bic <- plot_diagnostic_curve(
    diag_mclust_raw,
    y_col = "BIC",
    title_text = "Finite mixture model component-count profile",
    y_lab = "BIC",
    method_label = "Raw-indicator-based feature set"
  )

  p_mclust_raw_icl <- plot_diagnostic_curve(
    diag_mclust_raw,
    y_col = "ICL",
    title_text = "Finite mixture model component-count profile",
    y_lab = "ICL",
    method_label = "Raw-indicator-based feature set"
  )

  p_mclust_raw_sil <- plot_diagnostic_curve(
    diag_mclust_raw,
    y_col = "silhouette",
    title_text = "Finite mixture model secondary compactness summary",
    y_lab = "Average silhouette width",
    method_label = "Raw-indicator-based feature set"
  )

  (p_mclust_raw_bic | p_mclust_raw_icl) / p_mclust_raw_sil
}
```

```{r mclust-raw-fit}
if (isTRUE(params$use_mclust)) {
  raw_mclust_col <- pick_cluster_col("raw", "mclust", selected_k_raw)
  cluster_assignments[[raw_mclust_col]] <- fit_mclust_labels(raw_x, k = selected_k_raw, seed = params$seed)

  raw_mclust_features <- raw_feature_tbl %>%
    left_join(cluster_assignments %>% select(DataZone, all_of(raw_mclust_col)), by = "DataZone")

  kable(
    count_clusters(raw_mclust_features[[raw_mclust_col]]),
    caption = paste0("Cluster sizes for constrained comparator ", caption_safe(raw_mclust_col))
  ) %>%
    kable_styling(full_width = FALSE)
}
```

```{r mclust-raw-boxplot, fig.width=13, fig.height=9}
if (isTRUE(params$use_mclust)) {
  plot_domain_boxplots(
    raw_mclust_features,
    cluster_col = raw_mclust_col,
    title_text = paste0("Finite mixture comparator, raw feature set: ", raw_mclust_col, " (reported at k = ", selected_k_raw, ")")
  )
}
```

### PC-based clustering

```{r mclust-pc-diagnostics}
if (isTRUE(params$use_mclust)) {
  diag_mclust_pc <- method_diagnostics(pc_x, params$k_grid, method = "mclust", seed = params$seed) %>%
    mutate(feature_set = "pc")

  p_mclust_pc_bic <- plot_diagnostic_curve(
    diag_mclust_pc,
    y_col = "BIC",
    title_text = "Finite mixture model component-count profile",
    y_lab = "BIC",
    method_label = "PC-based feature set"
  )

  p_mclust_pc_icl <- plot_diagnostic_curve(
    diag_mclust_pc,
    y_col = "ICL",
    title_text = "Finite mixture model component-count profile",
    y_lab = "ICL",
    method_label = "PC-based feature set"
  )

  p_mclust_pc_sil <- plot_diagnostic_curve(
    diag_mclust_pc,
    y_col = "silhouette",
    title_text = "Finite mixture model secondary compactness summary",
    y_lab = "Average silhouette width",
    method_label = "PC-based feature set"
  )

  (p_mclust_pc_bic | p_mclust_pc_icl) / p_mclust_pc_sil
}
```

```{r mclust-pc-fit}
if (isTRUE(params$use_mclust)) {
  pc_mclust_col <- pick_cluster_col("pc", "mclust", selected_k_pc)
  cluster_assignments[[pc_mclust_col]] <- fit_mclust_labels(pc_x, k = selected_k_pc, seed = params$seed)

  pc_mclust_features <- pc_feature_tbl %>%
    left_join(cluster_assignments %>% select(DataZone, all_of(pc_mclust_col)), by = "DataZone")

  kable(
    count_clusters(pc_mclust_features[[pc_mclust_col]]),
    caption = paste0("Cluster sizes for constrained comparator ", caption_safe(pc_mclust_col))
  ) %>%
    kable_styling(full_width = FALSE)
}
```

```{r mclust-pc-boxplot, fig.width=13, fig.height=9}
if (isTRUE(params$use_mclust)) {
  plot_domain_boxplots(
    pc_mclust_features,
    cluster_col = pc_mclust_col,
    title_text = paste0("Finite mixture comparator, PC feature set: ", pc_mclust_col, " (reported at k = ", selected_k_pc, ")")
  )
}
```

### Comparator summary

```{r mclust-comparison}
if (isTRUE(params$use_mclust)) {
  mclust_comparison_tbl <- tibble(
    comparison = c("raw", "pc"),
    reporting_k = c(selected_k_raw, selected_k_pc),
    best_bic_k = c(
      best_metric_k(diag_mclust_raw, "BIC", "k"),
      best_metric_k(diag_mclust_pc, "BIC", "k")
    ),
    best_bic = c(
      best_metric_value(diag_mclust_raw, "BIC"),
      best_metric_value(diag_mclust_pc, "BIC")
    ),
    best_silhouette_k = c(
      best_metric_k(diag_mclust_raw, "silhouette", "k"),
      best_metric_k(diag_mclust_pc, "silhouette", "k")
    ),
    best_silhouette = c(
      best_metric_value(diag_mclust_raw, "silhouette"),
      best_metric_value(diag_mclust_pc, "silhouette")
    )
  )

  kable(
    mclust_comparison_tbl,
    digits = 3,
    caption = "Finite mixture comparator summary between raw and PC feature sets"
  ) %>%
    kable_styling(full_width = FALSE)
}
```

# Comparison of selected clustering results

This final comparison keeps the roles of the methods distinct: **K-means remains the headline partitioning result**, **C-means is an uncertainty overlay**, and **FMM is a secondary model-based comparator**.

## Diagnostic summaries

```{r selected-clustering-results}
selected_cluster_cols <- c(
  raw_kmeans_col,
  pc_kmeans_col,
  raw_hclust_col,
  pc_hclust_col
)

if (isTRUE(params$use_cmeans)) {
  selected_cluster_cols <- c(selected_cluster_cols, raw_cmeans_col, pc_cmeans_col)
}

if (isTRUE(params$use_mclust)) {
  selected_cluster_cols <- c(selected_cluster_cols, raw_mclust_col, pc_mclust_col)
}

selected_cluster_cols <- unique(selected_cluster_cols)
```

## Similarity cross-tables

```{r similarity}
similarity_pairs <- selected_similarity_pairs(selected_cluster_cols)

similarity_tbl <- compute_similarity_table(cluster_assignments, similarity_pairs)

kable(
  similarity_tbl %>% arrange(desc(ARI), desc(NMI)),
  digits = 3,
  caption = "Selected pairwise similarity metrics"
) %>%
  kable_styling(full_width = FALSE)
```

The square diagrams below are shown one metric at a time so that the labels and values remain readable in the knitted report.

```{r similarity-matrix-ari, fig.width=8.5, fig.height=7, out.width='88%'}
plot_similarity_heatmap(
  similarity_tbl,
  metric = "ARI",
  title_text = "ARI"
)
```

```{r similarity-matrix-nmi, fig.width=8.5, fig.height=7, out.width='88%'}
plot_similarity_heatmap(
  similarity_tbl,
  metric = "NMI",
  title_text = "NMI"
)
```

```{r similarity-matrix-vi, fig.width=8.5, fig.height=7, out.width='88%'}
plot_similarity_heatmap(
  similarity_tbl,
  metric = "VI",
  title_text = "VI"
)
```

## Primary combined diagnostics for the headline hard-partitioning methods

The combined comparison below is intentionally limited to **K-means** and **hierarchical clustering**, which carry the main hard-partition story. C-means and FMM are summarised in their own sections above; for those methods, silhouette and WSS are secondary diagnostics rather than the basis for the main conclusion.

```{r combined-diagnostics, fig.width=14, fig.height=12}
all_diagnostics <- dplyr::bind_rows(
  diag_kmeans_raw %>% dplyr::mutate(feature_set = "raw"),
  diag_kmeans_pc %>% dplyr::mutate(feature_set = "pc"),
  diag_hclust_raw %>% dplyr::mutate(feature_set = "raw"),
  diag_hclust_pc %>% dplyr::mutate(feature_set = "pc")
)

diag_methods <- all_diagnostics %>%
  mutate(
    method_label = dplyr::case_when(
      method == "kmeans" ~ "K-means",
      method == "hierarchical" ~ "Hierarchical",
      TRUE ~ method
    )
  )

method_levels <- diag_methods %>%
  dplyr::distinct(method, method_label) %>%
  dplyr::mutate(order = dplyr::case_when(
    method == "kmeans" ~ 1,
    method == "hierarchical" ~ 2,
    TRUE ~ 99
  )) %>%
  dplyr::arrange(order) %>%
  dplyr::pull(method_label)

diag_methods <- diag_methods %>%
  dplyr::mutate(method_label = factor(method_label, levels = method_levels))

plot_combined_diag <- function(data, y_col, title_text, y_lab, selected_k) {
  ggplot(data, aes(x = k, y = .data[[y_col]], colour = method_label)) +
    geom_vline(xintercept = selected_k, linetype = "dashed", colour = "grey40") +
    geom_line(linewidth = 0.9) +
    geom_point(size = 2) +
    annotate(
      "text",
      x = selected_k,
      y = Inf,
      label = paste0("chosen k = ", selected_k),
      vjust = 1.5,
      hjust = -0.05,
      size = 3.4,
      colour = "grey30"
    ) +
    scale_x_continuous(breaks = params$k_grid) +
    labs(
      title = title_text,
      x = "Number of clusters",
      y = y_lab,
      colour = "Method"
    ) +
    theme_minimal(base_size = 12) +
    theme(
      plot.title = element_text(face = "bold"),
      legend.position = "right"
    )
}

p_elbow_raw <- plot_combined_diag(
  diag_methods %>% dplyr::filter(feature_set == "raw"),
  y_col = "wss",
  title_text = "Combined elbow plot: raw-indicator-based feature set",
  y_lab = "Within-cluster sum of squares",
  selected_k = selected_k_raw
)

p_elbow_pc <- plot_combined_diag(
  diag_methods %>% dplyr::filter(feature_set == "pc"),
  y_col = "wss",
  title_text = "Combined elbow plot: PC-based feature set",
  y_lab = "Within-cluster sum of squares",
  selected_k = selected_k_pc
)

p_sil_raw <- plot_combined_diag(
  diag_methods %>% dplyr::filter(feature_set == "raw"),
  y_col = "silhouette",
  title_text = "Combined silhouette plot: raw-indicator-based feature set",
  y_lab = "Average silhouette width",
  selected_k = selected_k_raw
)

p_sil_pc <- plot_combined_diag(
  diag_methods %>% dplyr::filter(feature_set == "pc"),
  y_col = "silhouette",
  title_text = "Combined silhouette plot: PC-based feature set",
  y_lab = "Average silhouette width",
  selected_k = selected_k_pc
)

((p_elbow_raw | p_elbow_pc) / (p_sil_raw | p_sil_pc)) +
  patchwork::plot_layout(guides = "collect") &
  theme(legend.position = "right")
```


## External anchoring to overall SIMD rank

The two-panel figure below anchors the selected cluster solution back to overall SIMD position.
Panel (a) shows whether the selected clusters track overall deprivation rank.
Panel (b) uses a tightened five-indicator set and shows 95% confidence intervals around within-tertile cluster means, so that differences within broad overall-rank bands read more clearly as evidence rather than as an exploratory diagnostic.

```{r rank-anchor-panels, fig.width=12, fig.height=10.5, out.width='95%'}
headline_cluster_col <- raw_kmeans_col
rank_band_n <- 3L
rank_band_prefix <- "T"
rank_band_name_plural <- "tertiles"

if (is.null(overall_rank_source_col)) {
  cat("Overall SIMD rank column was not detected in the joined source data, so the external anchoring figure was skipped.")
} else if (!headline_cluster_col %in% names(cluster_assignments)) {
  stop(
    paste0(
      "The selected headline cluster column '", headline_cluster_col,
      "' was not found in cluster_assignments."
    ),
    call. = FALSE
  )
} else {
  indicator_label_map <- c(
    income_rate = "Income rate",
    employment_rate = "Employment rate",
    CIF = "CIF",
    no_qualifications = "No qualifications",
    DRUG = "DRUG"
  )

  requested_key_vars <- c(
    "income_rate",
    "employment_rate",
    "CIF",
    "no_qualifications",
    "DRUG"
  )

  missing_key_vars <- setdiff(requested_key_vars, names(board_complete))
  if (length(missing_key_vars) > 0) {
    stop(
      paste0(
        "The following requested panel (b) indicators were not found in board_complete: ",
        paste(missing_key_vars, collapse = ", ")
      ),
      call. = FALSE
    )
  }

  key_vars <- requested_key_vars

  rank_plot_df <- cluster_assignments %>%
    select(DataZone, cluster = all_of(headline_cluster_col)) %>%
    left_join(board_meta, by = "DataZone") %>%
    left_join(board_complete %>% select(DataZone, all_of(key_vars)), by = "DataZone") %>%
    filter(!is.na(overall_simd_rank)) %>%
    mutate(
      cluster = factor(cluster),
      overall_rank_band_num = dplyr::ntile(overall_simd_rank, rank_band_n),
      overall_rank_band = factor(
        overall_rank_band_num,
        levels = seq_len(rank_band_n),
        labels = paste0(rank_band_prefix, seq_len(rank_band_n))
      )
    )

  cluster_summary <- rank_plot_df %>%
    group_by(cluster) %>%
    summarise(median_rank = median(overall_simd_rank, na.rm = TRUE), .groups = "drop") %>%
    arrange(median_rank)

  cluster_levels_original <- levels(rank_plot_df$cluster)
  cluster_display_map <- setNames(
    paste0("Cluster ", cluster_levels_original),
    cluster_levels_original
  )

  if (length(cluster_levels_original) == 2L) {
    cluster_display_map[as.character(cluster_summary$cluster[1])] <- paste0(
      "Cluster ", cluster_summary$cluster[1], " (more deprived)"
    )
    cluster_display_map[as.character(cluster_summary$cluster[2])] <- paste0(
      "Cluster ", cluster_summary$cluster[2], " (less deprived)"
    )
  }

  cluster_display_levels <- unname(cluster_display_map[cluster_levels_original])
  cluster_palette_display <- setNames(
    cluster_palette[cluster_levels_original],
    cluster_display_levels
  )

  rank_plot_df <- rank_plot_df %>%
    mutate(
      cluster_display = factor(
        cluster_display_map[as.character(cluster)],
        levels = cluster_display_levels
      )
    )

  k_selected <- nlevels(rank_plot_df$cluster_display)
  max_rank_value <- max(rank_plot_df$overall_simd_rank, na.rm = TRUE)

  rank_breaks <- pretty(c(1, max_rank_value))
  rank_breaks <- rank_breaks[rank_breaks >= 1 & rank_breaks <= max_rank_value]

  p_rank <- ggplot(
    rank_plot_df,
    aes(x = cluster_display, y = overall_simd_rank, fill = cluster_display)
  ) +
    geom_violin(alpha = 0.25, width = 0.9, colour = NA, show.legend = FALSE) +
    geom_boxplot(width = 0.18, outlier.alpha = 0.20, show.legend = FALSE) +
    scale_fill_manual(values = cluster_palette_display, drop = FALSE, guide = "none") +
    scale_y_reverse(
      limits = c(max_rank_value, 1),
      breaks = rank_breaks,
      expand = expansion(mult = c(0.01, 0.01))
    ) +
    labs(
      title = "Overall SIMD rank by cluster",
      subtitle = paste0(
        params$target_board,
        " | selected raw-indicator k-means partition (k = ",
        k_selected,
        ")"
      ),
      x = NULL,
      y = "Overall SIMD rank (1 = most deprived)",
      fill = "Cluster"
    ) +
    theme_minimal(base_size = 12) +
    guides(fill = "none", colour = "none") +
    theme(
      plot.title = element_text(face = "bold"),
      axis.text.x = element_text(face = "bold"),
      legend.position = "none"
    )

  panel_b_df <- rank_plot_df %>%
    select(DataZone, cluster_display, overall_rank_band, all_of(key_vars)) %>%
    pivot_longer(
      cols = all_of(key_vars),
      names_to = "indicator",
      values_to = "value"
    ) %>%
    group_by(overall_rank_band, cluster_display, indicator) %>%
    summarise(
      n = sum(!is.na(value)),
      mean_value = mean(value, na.rm = TRUE),
      sd_value = sd(value, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    mutate(
      se_value = dplyr::if_else(n > 1, sd_value / sqrt(n), NA_real_),
      t_crit = dplyr::if_else(n > 1, qt(0.975, df = pmax(n - 1, 1)), NA_real_),
      ci_half = se_value * t_crit,
      ymin = mean_value - ci_half,
      ymax = mean_value + ci_half,
      indicator = factor(
        indicator,
        levels = key_vars,
        labels = unname(indicator_label_map[key_vars])
      )
    )

  panel_caption <- paste0(
    "Relationship between selected cluster membership and overall SIMD rank in ",
    params$target_board,
    ". Panel (a) shows the distribution of overall SIMD rank by cluster for the selected raw-indicator k-means partition (k = ",
    k_selected,
    "). Lower rank values indicate greater deprivation. Panel (b) shows mean indicator values by cluster within broad overall SIMD rank tertiles. ",
    "Error bars indicate 95% confidence intervals. Cluster-by-tertile representation is uneven, especially at the extremes, so those comparisons should be interpreted cautiously."
  )

  knitr::opts_current$set(fig.cap = panel_caption)

  p_band <- ggplot(
    panel_b_df,
    aes(x = overall_rank_band, y = mean_value, colour = cluster_display, group = cluster_display)
  ) +
    geom_errorbar(
      aes(ymin = ymin, ymax = ymax),
      width = 0.08,
      alpha = 0.6,
      linewidth = 0.65,
      na.rm = TRUE,
      show.legend = FALSE
    ) +
    geom_line(linewidth = 0.8, show.legend = TRUE) +
    geom_point(size = 2, show.legend = FALSE) +
    facet_wrap(~ indicator, scales = "free_y", ncol = 2) +
    scale_colour_manual(values = cluster_palette_display, drop = FALSE, name = "Cluster") +
    labs(
      title = "Domain composition within broad overall-rank bands",
      subtitle = paste0(
        toupper(substr(rank_band_name_plural, 1, 1)),
        substr(rank_band_name_plural, 2, nchar(rank_band_name_plural)),
        " of overall SIMD rank within ", params$target_board,
        " (T1 = most deprived)"
      ),
      x = "Overall SIMD rank band",
      y = "Mean indicator value",
      colour = "Cluster"
    ) +
    theme_minimal(base_size = 12) +
    theme(
      plot.title = element_text(face = "bold"),
      legend.position = "bottom",
      strip.text = element_text(face = "bold")
    )

  wrap_plots(p_rank, p_band, ncol = 1, heights = c(1, 1.55), guides = "collect") +
    plot_annotation(
      title = paste0(
        "Relationship between selected cluster membership and overall SIMD rank in ",
        params$target_board
      ),
      tag_levels = "a"
    ) &
    theme(legend.position = "bottom")
}
```

## Maps for selected clustering results

```{r maps, fig.width=13, fig.height=10}
map_sf <- board_sf_complete %>%
  left_join(cluster_assignments, by = "DataZone")

map_cols <- selected_cluster_cols[grepl("_k[0-9]+$", selected_cluster_cols)]

map_list <- purrr::map(
  map_cols,
  ~ plot_cluster_map(map_sf, .x, title_text = .x)
)

wrap_plots(map_list, ncol = 2) +
  plot_annotation(title = "Maps for selected clustering results")
```

## Difference maps

```{r difference-maps, fig.width=13, fig.height=10}
difference_plot_list <- list(
  plot_difference_map(map_sf, raw_kmeans_col, raw_hclust_col, "Raw: K-means vs hierarchical"),
  plot_difference_map(map_sf, pc_kmeans_col, pc_hclust_col, "PC: K-means vs hierarchical"),
  plot_difference_map(map_sf, raw_kmeans_col, pc_kmeans_col, "K-means: raw vs PC"),
  plot_difference_map(map_sf, raw_hclust_col, pc_hclust_col, "Hierarchical: raw vs PC")
)

if (isTRUE(params$use_cmeans)) {
  difference_plot_list <- c(
    difference_plot_list,
    list(
      plot_difference_map(map_sf, raw_cmeans_col, pc_cmeans_col, "C-means: raw vs PC")
    )
  )
}

if (isTRUE(params$use_mclust)) {
  difference_plot_list <- c(
    difference_plot_list,
    list(
      plot_difference_map(map_sf, raw_mclust_col, pc_mclust_col, "Finite mixture model: raw vs PC")
    )
  )
}

wrap_plots(difference_plot_list, ncol = 2) +
  plot_annotation(title = "Difference maps for selected clustering comparisons")
```

# Outputs passed to the vine document

```{r export-vine-inputs}
dir.create(params$output_dir, showWarnings = FALSE, recursive = TRUE)

readr::write_csv(board_complete, file.path(params$output_dir, paste0(output_prefix, "_analysis_base.csv")))
readr::write_csv(raw_feature_tbl, file.path(params$output_dir, paste0(output_prefix, "_feature_set_raw.csv")))
readr::write_csv(pc_feature_tbl, file.path(params$output_dir, paste0(output_prefix, "_feature_set_pc.csv")))
readr::write_csv(cluster_assignments, file.path(params$output_dir, paste0(output_prefix, "_cluster_assignments.csv")))

if (isTRUE(params$use_cmeans)) {
  readr::write_csv(
    raw_cmeans_membership_export,
    file.path(params$output_dir, paste0(output_prefix, "_cmeans_memberships_raw.csv"))
  )
  readr::write_csv(
    pc_cmeans_membership_export,
    file.path(params$output_dir, paste0(output_prefix, "_cmeans_memberships_pc.csv"))
  )
}
```

The next document should read:

- `r paste0(output_prefix, "_analysis_base.csv")`
- `r paste0(output_prefix, "_feature_set_raw.csv")`
- `r paste0(output_prefix, "_feature_set_pc.csv")`
- `r paste0(output_prefix, "_cluster_assignments.csv")`

If a reference C-means solution is carried forward as an uncertainty-aware comparator, it should also read:

- `r paste0(output_prefix, "_cmeans_memberships_raw.csv")`
- `r paste0(output_prefix, "_cmeans_memberships_pc.csv")`

These C-means membership files contain:

- `DataZone`
- `cluster_1_membership`
- `cluster_2_membership`
- `max_membership`
- `hard_label`
- `core_borderline_flag`

The chosen clustering result for vine fitting is then selected by its column name, for example:

- `raw_kmeans_k2`
- `pc_hclust_k2`
- `raw_cmeans_k2`
- `pc_mclust_k2`

A useful next-stage sensitivity check is to fit vines on **all members** first and then refit on a stricter **core-only subset** defined from `max_membership` (for example `r paste0(">= ", params$vine_core_membership_threshold)`) for the reference C-means solution. The updated vine document supports this directly through optional C-means membership inputs and a core-membership threshold parameter.

# Session info

```{r session-info}
sessionInfo()
```
