---
title: "Cluster-specific vine-copula modelling (Glasgow)"
author: "Guanyu"
date: "`r Sys.Date()`"
output:
  pdf_document:
    latex_engine: xelatex
  html_document:
    toc: true
    toc_float: true
    number_sections: true
params:
  output_dir: "outputs/glasgow"
  vine_dataset: "raw"
  cluster_column: "raw_kmeans_k2"
  cmeans_membership_source: !r NULL
  core_threshold_candidates: !r c(0.8, 0.7, 0.6)
  core_min_complete_n: 50
  vine_cores: 4
  selectioncrit: "BIC"
  treecrit: "tau"
  indeptest: true
  indep_level: 0.05
  trunclevel: !r NA
  familyset: !r NA
  compare_trees: !r 1:2
  node_label: "NO"
  condition_label: "NO"
  min_abs_tau: 0.00
  digits: 2
  seed: 123
  main_model_type: "RVine"
---

# Overview

This document:

1. fits **CVine** and **RVine** models within each cluster,
2. compares the two vine types in each cluster using **logLik, AIC, and BIC**,
3. treats **RVine** as the main model for interpretation,
4. compares **RVine Tree 1** and **RVine Tree 2** across the two clusters using:
   - individual tree plots,
   - an individual-tree table for cluster 1,
   - an individual-tree table for cluster 2,
   - a comparison table across clusters,
5. adds a **core-member robustness check** using the exported **reference C-means membership file**, choosing the strictest practical threshold from a candidate set,
6. only reports **core-only vine fits** when both clusters retain enough complete observations at the chosen threshold,
7. provides **CVine tree plots in the appendix**.

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  out.width = "100%",
  fig.width = 10,
  fig.height = 6,
  dpi = 300
)
```

```{r packages, include=FALSE}
required_pkgs <- c(
  "tidyverse",
  "VineCopula",
  "igraph",
  "ggplot2",
  "grid",
  "knitr",
  "kableExtra"
)
missing_pkgs <- setdiff(required_pkgs, rownames(installed.packages()))
if (length(missing_pkgs) > 0) {
  stop("Install these packages before knitting: ", paste(missing_pkgs, collapse = ", "))
}

suppressPackageStartupMessages({
  library(tidyverse)
  library(VineCopula)
  library(igraph)
  library(ggplot2)
  library(grid)
  library(knitr)
  library(kableExtra)
})
set.seed(params$seed)
```

# Helper functions

```{r helpers, include=FALSE}
`%||%` <- function(x, y) if (is.null(x)) y else x

assert_file_exists <- function(path) {
  if (!file.exists(path)) stop("File not found: ", path)
  invisible(path)
}

read_analysis_inputs <- function(output_dir) {
  base_path <- file.path(output_dir, "glasgow_analysis_base.csv")
  raw_path <- file.path(output_dir, "glasgow_feature_set_raw.csv")
  pc_path <- file.path(output_dir, "glasgow_feature_set_pc.csv")
  cluster_path <- file.path(output_dir, "glasgow_cluster_assignments.csv")
  cmeans_raw_path <- file.path(output_dir, "glasgow_cmeans_memberships_raw.csv")
  cmeans_pc_path <- file.path(output_dir, "glasgow_cmeans_memberships_pc.csv")
  purrr::walk(c(base_path, raw_path, pc_path, cluster_path), assert_file_exists)

  maybe_read_csv <- function(path) {
    if (!file.exists(path)) return(NULL)
    readr::read_csv(path, show_col_types = FALSE)
  }

  list(
    base = readr::read_csv(base_path, show_col_types = FALSE),
    raw = readr::read_csv(raw_path, show_col_types = FALSE),
    pc = readr::read_csv(pc_path, show_col_types = FALSE),
    clusters = readr::read_csv(cluster_path, show_col_types = FALSE),
    cmeans_raw = maybe_read_csv(cmeans_raw_path),
    cmeans_pc = maybe_read_csv(cmeans_pc_path)
  )
}

make_U <- function(X) {
  U <- VineCopula::pobs(as.matrix(X))
  colnames(U) <- colnames(X)
  U
}

as_rvine_matrix <- function(x) {
  if (inherits(x, "RVineMatrix")) return(x)
  if (is.list(x) && !is.null(x$RVM) && inherits(x$RVM, "RVineMatrix")) return(x$RVM)
  stop("`x` must be an `RVineMatrix` or a list containing `$RVM`.")
}

safe_cores <- function(x) {
  out <- suppressWarnings(as.integer(x[[1]]))
  if (length(out) == 0 || is.na(out) || out < 1) return(1L)
  out
}

is_informative_numeric <- function(x) {
  x <- x[is.finite(x)]
  if (length(x) < 2) return(FALSE)
  stats::sd(x) > 0 && dplyr::n_distinct(x) > 1
}

prepare_cluster_matrix <- function(df,
                                   cluster_value,
                                   cluster_col = "cluster",
                                   feature_cols = NULL) {
  dat <- df %>% filter(.data[[cluster_col]] == cluster_value)

  if (is.null(feature_cols)) {
    feature_cols <- setdiff(names(dat), c("DataZone", cluster_col, "max_membership"))
  }
  feature_cols <- intersect(feature_cols, names(dat))

  Xk <- dat %>%
    select(any_of(feature_cols)) %>%
    select(where(is.numeric)) %>%
    drop_na() %>%
    as.data.frame()

  if (ncol(Xk) == 0) {
    return(list(
      X = Xk,
      n_complete = nrow(Xk),
      d_complete = ncol(Xk),
      dropped_constant = character(),
      kept_features = character()
    ))
  }

  informative <- vapply(Xk, is_informative_numeric, logical(1))
  dropped_constant <- names(Xk)[!informative]
  Xk <- Xk[, informative, drop = FALSE]

  list(
    X = Xk,
    n_complete = nrow(Xk),
    d_complete = ncol(Xk),
    dropped_constant = dropped_constant,
    kept_features = names(Xk)
  )
}

fit_one_cluster_vine <- function(df,
                                 cluster_value,
                                 cluster_col = "cluster",
                                 feature_cols = NULL,
                                 type = c("RVine", "CVine"),
                                 selectioncrit = "BIC",
                                 treecrit = "tau",
                                 familyset = NA,
                                 trunclevel = NA,
                                 indeptest = TRUE,
                                 level = 0.05,
                                 cores = 4,
                                 progress = FALSE) {
  type <- match.arg(type)
  prep <- prepare_cluster_matrix(
    df = df,
    cluster_value = cluster_value,
    cluster_col = cluster_col,
    feature_cols = feature_cols
  )
  Xk <- prep$X

  if (prep$n_complete < 5) {
    stop("Cluster ", cluster_value, " has too few complete observations after filtering (n=", prep$n_complete, ").")
  }
  if (prep$d_complete < 2) {
    stop("Cluster ", cluster_value, " has too few informative variables for vine fitting (d=", prep$d_complete, ").")
  }

  Uk <- make_U(Xk)

  fit_once <- function(cores_use) {
    VineCopula::RVineStructureSelect(
      data = Uk,
      familyset = familyset,
      type = type,
      selectioncrit = selectioncrit,
      indeptest = indeptest,
      level = level,
      trunclevel = trunclevel,
      treecrit = treecrit,
      cores = cores_use,
      progress = progress
    )
  }

  requested_cores <- safe_cores(cores)
  fit_try <- tryCatch(fit_once(requested_cores), error = function(e) e)
  if (inherits(fit_try, "error") && requested_cores > 1L) {
    fit_try <- tryCatch(fit_once(1L), error = function(e) e)
  }
  if (inherits(fit_try, "error")) {
    stop(
      "Cluster ", cluster_value, " ", type,
      " fit failed after preprocessing: ",
      conditionMessage(fit_try)
    )
  }

  RVM <- fit_try
  RVM$names <- colnames(Uk)
  list(
    cluster = cluster_value,
    type = type,
    X = Xk,
    U = Uk,
    RVM = RVM,
    dropped_constant = prep$dropped_constant
  )
}

fit_clusterwise_vines <- function(df,
                                  cluster_col = "cluster",
                                  feature_cols = NULL,
                                  type = c("RVine", "CVine"),
                                  selectioncrit = "BIC",
                                  treecrit = "tau",
                                  familyset = NA,
                                  trunclevel = NA,
                                  indeptest = TRUE,
                                  level = 0.05,
                                  cores = 4) {
  type <- match.arg(type)
  clusters <- sort(unique(df[[cluster_col]]))
  clusters <- clusters[!is.na(clusters)]
  fits <- purrr::map(clusters, ~ fit_one_cluster_vine(
    df = df,
    cluster_value = .x,
    cluster_col = cluster_col,
    feature_cols = feature_cols,
    type = type,
    selectioncrit = selectioncrit,
    treecrit = treecrit,
    familyset = familyset,
    trunclevel = trunclevel,
    indeptest = indeptest,
    level = level,
    cores = cores
  ))
  names(fits) <- paste0("cluster_", clusters)
  fits
}

fit_vine_suite <- function(df, cluster_col = "cluster", feature_cols = NULL) {
  list(
    CVine = fit_clusterwise_vines(
      df = df,
      cluster_col = cluster_col,
      feature_cols = feature_cols,
      type = "CVine",
      selectioncrit = params$selectioncrit,
      treecrit = params$treecrit,
      familyset = params$familyset,
      trunclevel = params$trunclevel,
      indeptest = params$indeptest,
      level = params$indep_level,
      cores = params$vine_cores
    ),
    RVine = fit_clusterwise_vines(
      df = df,
      cluster_col = cluster_col,
      feature_cols = feature_cols,
      type = "RVine",
      selectioncrit = params$selectioncrit,
      treecrit = params$treecrit,
      familyset = params$familyset,
      trunclevel = params$trunclevel,
      indeptest = params$indeptest,
      level = params$indep_level,
      cores = params$vine_cores
    )
  )
}

summarise_fit_eligibility <- function(df,
                                      cluster_col = "cluster",
                                      feature_cols = NULL,
                                      min_n = 5) {
  if (!(cluster_col %in% names(df))) {
    return(tibble(
      cluster = numeric(),
      n_complete = integer(),
      d_complete = integer(),
      dropped_constant = character(),
      eligible_for_vine = logical(),
      fit_status = character()
    ))
  }

  clusters <- sort(unique(df[[cluster_col]]))
  clusters <- clusters[!is.na(clusters)]

  if (length(clusters) == 0) {
    return(tibble(
      cluster = numeric(),
      n_complete = integer(),
      d_complete = integer(),
      dropped_constant = character(),
      eligible_for_vine = logical(),
      fit_status = character()
    ))
  }

  purrr::map_dfr(clusters, function(cl) {
    prep <- prepare_cluster_matrix(
      df = df,
      cluster_value = cl,
      cluster_col = cluster_col,
      feature_cols = feature_cols
    )

    eligible <- prep$n_complete >= min_n && prep$d_complete >= 2
    reason <- dplyr::case_when(
      prep$n_complete < min_n ~ paste0("too few complete observations (n=", prep$n_complete, ")"),
      prep$d_complete < 2 ~ paste0("too few informative variables after filtering (d=", prep$d_complete, ")"),
      TRUE ~ "eligible"
    )

    tibble(
      cluster = cl,
      n_complete = prep$n_complete,
      d_complete = prep$d_complete,
      dropped_constant = if (length(prep$dropped_constant) == 0) "" else paste(prep$dropped_constant, collapse = ", "),
      eligible_for_vine = eligible,
      fit_status = reason
    )
  })
}

detect_cmeans_membership_source <- function(vine_dataset,
                                           cluster_column,
                                           requested_source = NULL) {
  if (!is.null(requested_source) && !is.na(requested_source) && nzchar(requested_source)) {
    return(requested_source)
  }
  if (!is.null(vine_dataset) && vine_dataset %in% c("raw", "pc")) {
    return(vine_dataset)
  }
  if (grepl("^raw_", cluster_column)) return("raw")
  if (grepl("^pc_", cluster_column)) return("pc")
  NULL
}

get_cmeans_membership_tbl <- function(inputs, source = NULL) {
  if (is.null(source) || is.na(source) || !nzchar(source)) return(NULL)
  switch(
    source,
    raw = inputs$cmeans_raw,
    pc = inputs$cmeans_pc,
    stop("Unknown C-means membership source: ", source)
  )
}

summarise_core_filter <- function(all_df, core_df, cluster_col = "cluster") {
  all_counts <- all_df %>%
    count(.data[[cluster_col]], name = "n_all") %>%
    rename(cluster = .data[[cluster_col]])

  core_counts <- core_df %>%
    count(.data[[cluster_col]], name = "n_core") %>%
    rename(cluster = .data[[cluster_col]])

  all_counts %>%
    full_join(core_counts, by = "cluster") %>%
    mutate(
      n_all = replace_na(n_all, 0L),
      n_core = replace_na(n_core, 0L),
      retained_pct = if_else(n_all > 0, n_core / n_all, NA_real_)
    ) %>%
    arrange(cluster)
}

safe_num <- function(x) {
  if (length(x) == 0 || is.null(x) || all(is.na(x))) return(NA_real_)
  as.numeric(x[1])
}


coerce_threshold_candidates <- function(x) {
  out <- suppressWarnings(as.numeric(x))
  out <- out[is.finite(out)]
  out <- unique(out)
  out <- out[out >= 0 & out <= 1]
  sort(out, decreasing = TRUE)
}

summarise_core_threshold_candidates <- function(all_df,
                                                thresholds,
                                                cluster_col = "cluster",
                                                feature_cols = NULL,
                                                min_n = 50) {
  clusters_all <- sort(unique(all_df[[cluster_col]]))
  clusters_all <- clusters_all[!is.na(clusters_all)]

  purrr::map_dfr(thresholds, function(thr) {
    core_df <- all_df %>%
      filter(!is.na(max_membership) & max_membership >= thr)

    counts_tbl <- summarise_core_filter(all_df, core_df, cluster_col = cluster_col)
    elig_tbl <- summarise_fit_eligibility(
      core_df,
      cluster_col = cluster_col,
      feature_cols = feature_cols,
      min_n = min_n
    )

    counts_tbl %>%
      full_join(elig_tbl, by = "cluster") %>%
      complete(cluster = clusters_all) %>%
      mutate(
        threshold = thr,
        n_all = replace_na(n_all, 0L),
        n_core = replace_na(n_core, 0L),
        retained_pct = if_else(n_all > 0, n_core / n_all, NA_real_),
        n_complete = replace_na(n_complete, 0L),
        d_complete = replace_na(d_complete, 0L),
        dropped_constant = replace_na(dropped_constant, ""),
        eligible_for_vine = replace_na(eligible_for_vine, FALSE),
        fit_status = replace_na(fit_status, paste0("too few complete observations (n=0; need at least ", min_n, ")"))
      ) %>%
      arrange(cluster)
  })
}

choose_core_threshold <- function(threshold_summary) {
  if (nrow(threshold_summary) == 0) return(NA_real_)

  threshold_status <- threshold_summary %>%
    group_by(threshold) %>%
    summarise(
      all_clusters_eligible = all(eligible_for_vine),
      .groups = "drop"
    ) %>%
    arrange(desc(threshold))

  chosen <- threshold_status %>%
    filter(all_clusters_eligible) %>%
    slice_head(n = 1) %>%
    pull(threshold)

  if (length(chosen) == 0) NA_real_ else chosen[[1]]
}

summarise_vine_fits <- function(fit_list) {
  if (length(fit_list) == 0) {
    return(tibble(
      type = character(),
      cluster = integer(),
      n = integer(),
      d = integer(),
      logLik = double(),
      AIC = double(),
      BIC = double()
    ))
  }

  purrr::imap_dfr(fit_list, function(obj, nm) {
    cluster_guess <- suppressWarnings(as.integer(gsub("^cluster_", "", nm)))
    ll <- safe_num(obj$RVM$logLik)
    if (is.na(ll)) ll <- tryCatch({
      out <- VineCopula::RVineLogLik(obj$U, obj$RVM)
      if (length(out) > 1) sum(out, na.rm = TRUE) else safe_num(out)
    }, error = function(e) NA_real_)
    aic <- safe_num(obj$RVM$AIC)
    if (is.na(aic) && !is.na(ll)) aic <- -2 * ll + 2 * sum(!is.na(obj$RVM$family))
    bic <- safe_num(obj$RVM$BIC)
    if (is.na(bic) && !is.na(ll)) bic <- -2 * ll + log(nrow(obj$U)) * sum(!is.na(obj$RVM$family))
    tibble(
      type = obj$type %||% NA_character_,
      cluster = obj$cluster %||% cluster_guess,
      n = nrow(obj$U),
      d = ncol(obj$U),
      logLik = ll,
      AIC = aic,
      BIC = bic
    )
  })
}

select_best_type_by_cluster <- function(fit_summary, metric = c("BIC", "AIC", "logLik")) {
  metric <- match.arg(metric)
  if (nrow(fit_summary) == 0) {
    return(fit_summary %>% mutate(selected = logical()))
  }
  fit_summary %>%
    group_by(cluster) %>%
    arrange(
      if (metric == "logLik") desc(.data[[metric]]) else .data[[metric]],
      BIC, AIC, desc(logLik),
      .by_group = TRUE
    ) %>%
    mutate(selected = row_number() == 1L) %>%
    ungroup()
}

format_cond_label <- function(cond_idx, var_names, mode = c("NO", "name")) {
  mode <- match.arg(mode)
  idx <- cond_idx[!is.na(cond_idx)]
  if (length(idx) == 0) return("")
  idx <- sort(unique(idx))
  if (mode == "NO") paste(idx, collapse = ",") else paste(var_names[idx], collapse = ", ")
}

extract_tree_edges <- function(RVM, tree = 1, var_names = RVM$names) {
  out <- capture.output(print(RVM))
  i1 <- grep(paste0("^Tree\\s+", tree, "\\b"), out)[1]
  i2 <- grep(paste0("^Tree\\s+", tree + 1, "\\b"), out)[1]
  if (is.na(i1)) stop("Could not find Tree ", tree, " in print(RVM).")
  if (is.na(i2)) i2 <- length(out) + 1
  lines <- out[(i1 + 1):(i2 - 1)]
  lines <- trimws(gsub("^#+\\s*", "", lines))
  lines <- lines[nzchar(lines)]
  lines <- lines[grepl("^[0-9]+\\s*,\\s*[0-9]+", lines)]
  if (length(lines) == 0) return(tibble())

  toks <- strsplit(lines, "\\s+")
  pair_tok <- vapply(toks, function(z) if (length(z) >= 1) z[1] else NA_character_, character(1))
  fam_raw <- vapply(toks, function(z) if (length(z) >= 2) z[2] else "Independence", character(1))
  tau <- suppressWarnings(as.numeric(stringr::str_match(lines, "tau\\s*=\\s*([-0-9.]+)")[, 2]))
  tau[is.na(tau)] <- 0

  pair_part <- sub("[;|].*$", "", pair_tok)
  cond_part <- ifelse(grepl("[;|]", pair_tok), sub("^[^;|]+[;|]", "", pair_tok), "")
  parts <- strsplit(pair_part, ",", fixed = TRUE)
  v1 <- trimws(vapply(parts, function(z) if (length(z) >= 1) z[1] else NA_character_, character(1)))
  v2 <- trimws(vapply(parts, function(z) if (length(z) >= 2) z[2] else NA_character_, character(1)))

  idx1 <- suppressWarnings(as.integer(v1))
  idx2 <- suppressWarnings(as.integer(v2))
  if (anyNA(idx1)) idx1[is.na(idx1)] <- match(v1[is.na(idx1)], var_names)
  if (anyNA(idx2)) idx2[is.na(idx2)] <- match(v2[is.na(idx2)], var_names)

  cond_idx <- lapply(cond_part, function(s) {
    s <- trimws(s)
    if (!nzchar(s)) return(integer(0))
    z <- trimws(unlist(strsplit(s, ",", fixed = TRUE)))
    zi <- suppressWarnings(as.integer(z))
    if (anyNA(zi)) zi[is.na(zi)] <- match(z[is.na(zi)], var_names)
    zi[!is.na(zi)]
  })

  tibble(
    idx1 = idx1,
    idx2 = idx2,
    pair_no = paste(pmin(idx1, idx2), pmax(idx1, idx2), sep = " -- "),
    pair_name = paste(var_names[pmin(idx1, idx2)], var_names[pmax(idx1, idx2)], sep = " — "),
    cond_idx = cond_idx,
    cond_no = vapply(cond_idx, format_cond_label, character(1), var_names = var_names, mode = "NO"),
    cond_name = vapply(cond_idx, format_cond_label, character(1), var_names = var_names, mode = "name"),
    family = fam_raw,
    tau = tau,
    abs_tau = abs(tau)
  ) %>% filter(!is.na(idx1), !is.na(idx2))
}


choose_tree_root <- function(RVM, tree = 1) {
  vn <- RVM$names %||% paste0("V", seq_len(nrow(RVM$Matrix)))
  nodes <- tibble(name = as.character(seq_along(vn)))
  ed <- extract_tree_edges(RVM, tree = tree, var_names = vn) %>%
    transmute(from = as.character(idx1), to = as.character(idx2)) %>%
    distinct()
  g <- igraph::graph_from_data_frame(ed, directed = FALSE, vertices = nodes)
  deg <- igraph::degree(g)
  igraph::V(g)$name[which.max(deg)]
}

make_tree_layout <- function(RVM, tree = 1, root = NULL,
                             orientation = c("auto", "vertical", "horizontal", "radial"),
                             seed = 1) {
  orientation <- match.arg(orientation)
  vn <- RVM$names %||% paste0("V", seq_len(nrow(RVM$Matrix)))
  nodes <- tibble(name = as.character(seq_along(vn)))

  ed <- extract_tree_edges(RVM, tree = tree, var_names = vn) %>%
    transmute(from = as.character(idx1), to = as.character(idx2)) %>%
    distinct() %>%
    arrange(as.integer(from), as.integer(to))

  g <- igraph::graph_from_data_frame(ed, directed = FALSE, vertices = nodes)
  node_names <- igraph::V(g)$name

  if (is.null(root) || !(as.character(root) %in% node_names)) {
    root <- choose_tree_root(RVM, tree = tree)
  }
  root_idx <- match(as.character(root), node_names)

  if (orientation == "auto") {
    orientation <- if (tree == 1) "radial" else "vertical"
  }

  if (orientation == "radial") {
    xy <- igraph::layout_as_tree(g, root = root_idx, circular = TRUE)
    coords <- tibble(name = node_names, x = xy[, 1], y = xy[, 2])
  } else if (igraph::is_tree(g)) {
    xy <- igraph::layout_as_tree(g, root = root_idx, circular = FALSE)
    coords <- tibble(name = node_names, x = -xy[, 2], y = xy[, 1])
    if (orientation == "horizontal") {
      coords <- coords %>% transmute(name, x = y, y = x)
    }
  } else {
    set.seed(seed)
    xy <- igraph::layout_with_fr(g)
    coords <- tibble(name = node_names, x = xy[, 1], y = xy[, 2])
    if (orientation == "horizontal") {
      coords <- coords %>% transmute(name, x = y, y = x)
    }
  }
  coords
}

plot_individual_tree <- function(fit,
                                 tree = 1,
                                 node_label = c("NO", "name"),
                                 condition_label = c("NO", "name"),
                                 edge_label = c("auto", "tau", "cond", "tau+cond", "none"),
                                 min_abs_tau = NULL,
                                 digits = 2,
                                 orientation = c("auto", "vertical", "horizontal", "radial"),
                                 show_indep = TRUE,
                                 title = NULL) {
  node_label <- match.arg(node_label)
  condition_label <- match.arg(condition_label)
  edge_label <- match.arg(edge_label)
  orientation <- match.arg(orientation)
  RVM <- as_rvine_matrix(fit)
  var_names <- RVM$names %||% paste0("V", seq_len(nrow(RVM$Matrix)))

  if (edge_label == "auto") edge_label <- if (tree == 1) "tau" else "tau+cond"
  if (is.null(min_abs_tau)) min_abs_tau <- if (tree == 1) 0.20 else 0.10

  nodes <- tibble(
    name = as.character(seq_along(var_names)),
    node_text = if (node_label == "NO") as.character(seq_along(var_names)) else var_names
  )

  ed <- extract_tree_edges(RVM, tree = tree, var_names = var_names) %>%
    mutate(
      from = as.character(idx1),
      to = as.character(idx2),
      tau_label = formatC(tau, format = "f", digits = digits),
      cond_label = if (condition_label == "NO") {
        ifelse(cond_no == "", "", paste0("| ", cond_no))
      } else {
        ifelse(cond_name == "", "", paste0("| ", cond_name))
      },
      is_indep = abs_tau < 1e-12,
      label = dplyr::case_when(
        edge_label == "none" ~ "",
        edge_label == "tau" ~ dplyr::if_else(abs_tau >= min_abs_tau & !is_indep, tau_label, ""),
        edge_label == "cond" ~ cond_label,
        edge_label == "tau+cond" ~ dplyr::case_when(
          abs_tau >= min_abs_tau & !is_indep & cond_label != "" ~ paste(tau_label, cond_label),
          abs_tau >= min_abs_tau & !is_indep & cond_label == "" ~ tau_label,
          TRUE ~ cond_label
        ),
        TRUE ~ ""
      )
    )

  coords <- make_tree_layout(RVM, tree = tree, orientation = orientation)

  edf <- ed %>% mutate(
    x = coords$x[match(from, coords$name)],
    y = coords$y[match(from, coords$name)],
    xend = coords$x[match(to, coords$name)],
    yend = coords$y[match(to, coords$name)],
    xm = (x + xend) / 2,
    ym = (y + yend) / 2
  )

  ggplot() +
    { if (show_indep) geom_segment(
      data = edf %>% filter(is_indep),
      aes(x = x, y = y, xend = xend, yend = yend),
      colour = "grey80", linetype = "dashed", linewidth = 0.6, alpha = 0.9
    ) } +
    geom_segment(
      data = edf %>% filter(!is_indep | !show_indep),
      aes(x = x, y = y, xend = xend, yend = yend, colour = family),
      linewidth = 0.9, alpha = 0.95
    ) +
    geom_label(
      data = edf %>% filter(label != ""),
      aes(x = xm, y = ym, label = label),
      size = 2.5, label.size = 0, fill = "white", alpha = 0.82,
      label.padding = unit(0.05, "lines")
    ) +
    geom_text(
      data = coords %>% left_join(nodes, by = "name"),
      aes(x = x, y = y, label = node_text),
      size = 4.0, fontface = "bold"
    ) +
    coord_equal(clip = "off") +
    scale_x_continuous(expand = expansion(mult = 0.08)) +
    scale_y_continuous(expand = expansion(mult = 0.10)) +
    theme_void(base_size = 12) +
    labs(title = title %||% paste("Tree", tree), colour = "Copula family") +
    theme(
      plot.title = element_text(face = "bold", hjust = 0.5),
      legend.position = "bottom"
    )
}

make_individual_tree_table <- function(fit, tree = 1, condition_label = c("NO", "name"), digits = 2) {
  condition_label <- match.arg(condition_label)
  RVM <- as_rvine_matrix(fit)
  var_names <- RVM$names

  out <- extract_tree_edges(RVM, tree = tree, var_names = var_names) %>%
    transmute(
      pair = pair_name,
      condition = if (condition_label == "NO") cond_no else cond_name,
      family = family,
      tau = round(tau, digits),
    )

  if (tree == 1) {
    out <- out %>% select(-condition)
  }

  out
}

make_compare_trees_table <- function(fit_a,
                                     fit_b,
                                     tree = 1,
                                     label_a = "Cluster 1",
                                     label_b = "Cluster 2",
                                     condition_label = c("NO", "name"),
                                     digits = 2) {
  condition_label <- match.arg(condition_label)
  RVM_a <- as_rvine_matrix(fit_a)
  RVM_b <- as_rvine_matrix(fit_b)
  short_a <- paste0("c", fit_a$cluster %||% label_a)
  short_b <- paste0("c", fit_b$cluster %||% label_b)

  ea <- extract_tree_edges(RVM_a, tree = tree, var_names = RVM_a$names) %>%
    mutate(edge_id = paste(pair_name, ifelse(cond_name == "", "", paste0(" | ", cond_name))))
  eb <- extract_tree_edges(RVM_b, tree = tree, var_names = RVM_b$names) %>%
    mutate(edge_id = paste(pair_name, ifelse(cond_name == "", "", paste0(" | ", cond_name))))

  out <- full_join(
    ea %>% select(edge_id, pair_name, cond_no, cond_name, family_a = family, tau_a = tau),
    eb %>% select(edge_id, pair_name, cond_no, cond_name, family_b = family, tau_b = tau),
    by = c("edge_id", "pair_name", "cond_no", "cond_name")
  ) %>%
    mutate(
      across(c(tau_a, tau_b), ~ replace_na(., 0)),
      family_a = replace_na(family_a, "—"),
      family_b = replace_na(family_b, "—"),
      abs_delta_tau = abs(tau_a - tau_b)
    ) %>%
    transmute(
      pair = pair_name,
      condition = if (condition_label == "NO") cond_no else cond_name,
      !!paste0("family ", short_a) := family_a,
      !!paste0("family ", short_b) := family_b,
      !!paste0("tau ", short_a) := round(tau_a, digits),
      !!paste0("tau ", short_b) := round(tau_b, digits),
      `abs delta tau` = round(abs_delta_tau, digits)
    ) %>%
    arrange(desc(`abs delta tau`), pair)

  if (tree == 1) {
    out <- out %>% select(-condition)
  }

  out
}

latex_safe_caption <- function(x) {
  x <- gsub("_", " ", x, fixed = TRUE)
  x <- gsub("|", " ", x, fixed = TRUE)
  x
}

report_table <- function(x, caption, digits = 2, long = NULL) {
  cap <- latex_safe_caption(caption)
  if (is.null(long)) long <- ncol(x) > 7 || nrow(x) > 20
  if (knitr::is_latex_output()) {
    kable(x,
          format = "latex",
          booktabs = TRUE,
          longtable = long,
          caption = cap,
          digits = digits,
          escape = TRUE) %>%
      kable_styling(
        latex_options = c("hold_position", if (long) "repeat_header"),
        full_width = FALSE,
        font_size = 9
      )
  } else {
    kable(x, digits = digits, caption = cap) %>%
      kable_styling(full_width = FALSE)
  }
}
```

# Read inputs and build vine dataset

```{r data, include=FALSE}
inputs <- read_analysis_inputs(params$output_dir)
analysis_tbl <- switch(
  params$vine_dataset,
  base = inputs$base,
  raw = inputs$raw,
  pc = inputs$pc,
  stop("Unknown `params$vine_dataset`: ", params$vine_dataset)
)

cluster_tbl <- inputs$clusters %>% select(DataZone, cluster = all_of(params$cluster_column))
membership_source <- detect_cmeans_membership_source(
  vine_dataset = params$vine_dataset,
  cluster_column = params$cluster_column,
  requested_source = params$cmeans_membership_source
)
if (is.null(membership_source)) {
  stop("Could not determine the reference C-means membership source. Set `params$cmeans_membership_source` to 'raw' or 'pc'.")
}

cmeans_membership_tbl <- get_cmeans_membership_tbl(inputs, membership_source)
if (is.null(cmeans_membership_tbl)) {
  stop(
    "Missing exported reference C-means membership file for source: ", membership_source,
    ". Expected one of `glasgow_cmeans_memberships_raw.csv` or `glasgow_cmeans_memberships_pc.csv` in `params$output_dir`."
  )
}
if (!("max_membership" %in% names(cmeans_membership_tbl))) {
  stop("The reference C-means membership file must contain a `max_membership` column.")
}

feature_cols <- setdiff(names(analysis_tbl), "DataZone")

vine_df <- analysis_tbl %>%
  select(DataZone, all_of(feature_cols)) %>%
  left_join(cluster_tbl, by = "DataZone") %>%
  left_join(cmeans_membership_tbl %>% select(DataZone, max_membership), by = "DataZone")

full_fit_eligibility <- summarise_fit_eligibility(
  vine_df,
  cluster_col = "cluster",
  feature_cols = feature_cols,
  min_n = 5
)

core_threshold_candidates <- coerce_threshold_candidates(params$core_threshold_candidates)
if (length(core_threshold_candidates) == 0) {
  stop("`params$core_threshold_candidates` must contain at least one numeric threshold between 0 and 1.")
}

core_threshold_summary <- summarise_core_threshold_candidates(
  all_df = vine_df,
  thresholds = core_threshold_candidates,
  cluster_col = "cluster",
  feature_cols = feature_cols,
  min_n = params$core_min_complete_n
)
selected_core_threshold <- choose_core_threshold(core_threshold_summary)

if (is.na(selected_core_threshold)) {
  vine_df_core <- NULL
  core_fit_eligibility <- core_threshold_summary %>%
    filter(threshold == min(core_threshold_candidates))
  eligible_core_clusters <- integer(0)
} else {
  vine_df_core <- vine_df %>%
    filter(!is.na(max_membership) & max_membership >= selected_core_threshold)

  core_fit_eligibility <- core_threshold_summary %>%
    filter(threshold == selected_core_threshold)

  eligible_core_clusters <- core_fit_eligibility %>%
    filter(eligible_for_vine) %>%
    pull(cluster)
}
```

# Fit CVine and RVine in each cluster

```{r fit-models, include=FALSE}
ineligible_full_clusters <- full_fit_eligibility %>%
  filter(!eligible_for_vine) %>%
  pull(cluster)
if (length(ineligible_full_clusters) > 0) {
  stop(
    "These full-sample clusters are not eligible for vine fitting: ",
    paste(ineligible_full_clusters, collapse = ", "),
    ". Check the eligibility table inputs and missing-data pattern."
  )
}

vine_fit_list <- fit_vine_suite(
  df = vine_df,
  cluster_col = "cluster",
  feature_cols = feature_cols
)

fit_summary <- bind_rows(
  summarise_vine_fits(vine_fit_list$CVine),
  summarise_vine_fits(vine_fit_list$RVine)
) %>% arrange(cluster, type)

if (!is.null(vine_df_core) && length(eligible_core_clusters) > 0) {
  core_fit_input <- vine_df_core %>% filter(cluster %in% eligible_core_clusters)
  core_fit_list <- fit_vine_suite(
    df = core_fit_input,
    cluster_col = "cluster",
    feature_cols = feature_cols
  )
  core_fit_summary <- bind_rows(
    summarise_vine_fits(core_fit_list$CVine),
    summarise_vine_fits(core_fit_list$RVine)
  ) %>% arrange(cluster, type)
} else {
  core_fit_list <- NULL
  core_fit_summary <- tibble(
    type = character(),
    cluster = integer(),
    n = integer(),
    d = integer(),
    logLik = double(),
    AIC = double(),
    BIC = double()
  )
}
```

# 1. Full-sample CVine versus RVine model comparison by cluster

```{r full-eligibility}
report_table(full_fit_eligibility, "Full-sample vine-fit eligibility by cluster", digits = 2)
```

```{r fit-summary}
report_table(fit_summary, "CVine and RVine fit statistics by cluster")
```

```{r selected-models}
selected_models <- select_best_type_by_cluster(fit_summary, metric = params$selectioncrit)
model_choice_tbl <- selected_models %>%
  select(cluster, type, logLik, AIC, BIC, selected) %>%
  arrange(cluster, desc(selected), BIC, AIC, desc(logLik))
report_table(model_choice_tbl, paste0("Model choice by cluster using ", params$selectioncrit))
```

```{r full-sample-model-choice-text, results='asis'}
full_bic_choice <- select_best_type_by_cluster(fit_summary, metric = "BIC") %>%
  filter(selected)

if (nrow(full_bic_choice) > 0 && all(full_bic_choice$type == "RVine")) {
  cat("In the full sample, BIC selects RVine in both clusters.")
} else {
  full_bic_sentence <- full_bic_choice %>%
    transmute(txt = paste0("cluster ", cluster, ": ", type)) %>%
    pull(txt) %>%
    paste(collapse = "; ")

  cat("In the full sample, BIC selects ", full_bic_sentence, ".")
}
```

# 2. Core-member robustness check using the reference C-means membership file

This section treats the core-member repeat as a **robustness check**, not as the main result. The threshold is chosen from the candidate set `r paste(core_threshold_candidates, collapse = ", ")` using a practical rule: retain the **strictest** threshold that still leaves at least `r params$core_min_complete_n` complete observations in **both** clusters. If no candidate threshold meets that requirement, the document reports the threshold summary but does not fit core-only vine models.

```{r core-threshold-summary}
core_threshold_tbl_display <- core_threshold_summary %>%
  transmute(
    threshold,
    cluster,
    n_all,
    n_core,
    kept_pct = 100 * retained_pct,
    status = if_else(eligible_for_vine, "eligible", fit_status)
  )

report_table(
  core_threshold_tbl_display,
  paste0(
    "Core-member threshold summary from the ",
    membership_source,
    " reference C-means file"
  ),
  digits = 1,
  long = FALSE
)
```

```{r core-threshold-decision, results='asis'}
if (is.na(selected_core_threshold)) {
  cat(
    "No candidate threshold preserved at least ", params$core_min_complete_n,
    " complete observations in both clusters. The full-sample vine analysis remains the main result, and no core-only vine fits are reported."
  )
} else {
  highest_candidate <- max(core_threshold_candidates, na.rm = TRUE)
  highest_candidate_ok <- core_threshold_summary %>%
    filter(threshold == highest_candidate) %>%
    summarise(all_ok = all(replace_na(eligible_for_vine, FALSE))) %>%
    pull(all_ok)

  if (!isTRUE(highest_candidate_ok) && selected_core_threshold < highest_candidate) {
    cat(
      "This confirms that a fixed threshold of ", highest_candidate,
      " would be too severe for Glasgow, so the robustness check is conducted at ",
      selected_core_threshold,
      ", the strictest threshold that preserves adequate sample size in both clusters."
    )
  } else {
    cat(
      "The selected core-member threshold is ", selected_core_threshold,
      ", which is the strictest candidate that preserves at least ", params$core_min_complete_n,
      " complete observations in both clusters."
    )
  }
}
```

```{r core-model-comparison-table}
if (is.na(selected_core_threshold) || nrow(core_fit_summary) == 0) {
  cat(
    "Core-member vine fits were treated as a robustness check and were only attempted at thresholds that preserved adequate complete-case sample size in both clusters."
  )
} else {
  core_bic_choice <- select_best_type_by_cluster(core_fit_summary, metric = "BIC") %>%
    filter(selected) %>%
    select(cluster, bic_selected_type = type)

  core_aic_choice <- select_best_type_by_cluster(core_fit_summary, metric = "AIC") %>%
    filter(selected) %>%
    select(cluster, aic_selected_type = type)

  core_model_comparison_tbl <- core_fit_summary %>%
    left_join(core_bic_choice, by = "cluster") %>%
    left_join(core_aic_choice, by = "cluster") %>%
    mutate(
      selected_BIC = type == bic_selected_type,
      selected_AIC = type == aic_selected_type
    ) %>%
    select(cluster, type, n, d, logLik, AIC, BIC, selected_BIC, selected_AIC) %>%
    arrange(cluster, desc(selected_BIC), BIC, AIC, desc(logLik))

  report_table(
    core_model_comparison_tbl,
    paste0("Core-member CVine versus RVine comparison at threshold ", selected_core_threshold)
  )
}
```

```{r core-selected-models-text, results='asis'}
if (!is.na(selected_core_threshold) && nrow(core_fit_summary) > 0) {
  core_bic_choice <- select_best_type_by_cluster(core_fit_summary, metric = "BIC") %>%
    filter(selected)
  core_aic_choice <- select_best_type_by_cluster(core_fit_summary, metric = "AIC") %>%
    filter(selected)

  bic_all_rvine <- nrow(core_bic_choice) > 0 && all(core_bic_choice$type == "RVine")
  aic_cluster1_cvine <- core_aic_choice %>%
    filter(cluster == 1) %>%
    summarise(is_cvine = dplyr::first(type) == "CVine") %>%
    pull(is_cvine)

  if (isTRUE(bic_all_rvine) && isTRUE(aic_cluster1_cvine)) {
    cat(
      "In the core-member robustness sample at threshold ", selected_core_threshold,
      ", BIC again selects RVine in both clusters, while AIC is more mixed in cluster 1."
    )
  } else {
    bic_sentence <- core_bic_choice %>%
      transmute(txt = paste0("cluster ", cluster, ": ", type)) %>%
      pull(txt) %>%
      paste(collapse = "; ")
    aic_sentence <- core_aic_choice %>%
      transmute(txt = paste0("cluster ", cluster, ": ", type)) %>%
      pull(txt) %>%
      paste(collapse = "; ")

    cat(
      "In the core-member robustness sample at threshold ", selected_core_threshold,
      ", BIC selects ", bic_sentence,
      ". Using AIC, the preferred model is ", aic_sentence, "."
    )
  }
}
```

# 3. RVine interpretation: compare Trees 1 and 2 across clusters

We treat **RVine** as the main model for interpretation and compare the fitted trees across the two clusters. The full-sample comparison is the headline result; the core-member repeat above is only a robustness check when a practical threshold leaves enough complete observations in both clusters.

```{r rvine-objects, include=FALSE}
rvine_fits <- vine_fit_list[[params$main_model_type]]
clusters_available <- names(rvine_fits)
if (length(clusters_available) < 2) stop("Need at least two fitted clusters for comparison.")
rv1 <- rvine_fits[[1]]
rv2 <- rvine_fits[[2]]
label_1 <- paste(params$main_model_type, "Cluster", rv1$cluster)
label_2 <- paste(params$main_model_type, "Cluster", rv2$cluster)
```

## Tree 1

### Tree plots

```{r rvine-tree1-plots, fig.height=10}
plot_individual_tree(rv1, tree = 1, node_label = params$node_label, condition_label = params$condition_label,
                     edge_label = "tau", min_abs_tau = params$min_abs_tau, digits = params$digits,
                     orientation = "radial", title = paste(label_1, "- Tree 1"))
plot_individual_tree(rv2, tree = 1, node_label = params$node_label, condition_label = params$condition_label,
                     edge_label = "tau", min_abs_tau = params$min_abs_tau, digits = params$digits,
                     orientation = "radial", title = paste(label_2, "- Tree 1"))
```

### Key structure table: cluster 1

```{r rvine-tree1-table1}
report_table(make_individual_tree_table(rv1, tree = 1, condition_label = "name", digits = params$digits),
           paste(label_1, "Tree 1"))
```

### Key structure table: cluster 2

```{r rvine-tree1-table2}
report_table(make_individual_tree_table(rv2, tree = 1, condition_label = "name", digits = params$digits),
           paste(label_2, "Tree 1"))

```

### Comparison table

```{r rvine-tree1-compare}
report_table(make_compare_trees_table(rv1, rv2, tree = 1, label_a = label_1, label_b = label_2,
                                    condition_label = "name", digits = params$digits),
           paste(params$main_model_type, "Tree 1 comparison across clusters"))
```

```{r tree2-pagebreak, results="asis", echo=FALSE}
cat("\newpage")
```

## Tree 2

### Tree plots

```{r rvine-tree2-plots, fig.height=10}
plot_individual_tree(rv1, tree = 2, node_label = params$node_label, condition_label = params$condition_label,
                     edge_label = "tau+cond", min_abs_tau = params$min_abs_tau, digits = params$digits,
                     orientation = "vertical", title = paste(label_1, "- Tree 2"))
plot_individual_tree(rv2, tree = 2, node_label = params$node_label, condition_label = params$condition_label,
                     edge_label = "tau+cond", min_abs_tau = params$min_abs_tau, digits = params$digits,
                     orientation = "vertical", title = paste(label_2, "- Tree 2"))
```

### Key structure table: cluster 1

```{r rvine-tree2-table1}
report_table(make_individual_tree_table(rv1, tree = 2, condition_label = "name", digits = params$digits),
           paste(label_1, "Tree 2"))
```

### Key structure table: cluster 2

```{r rvine-tree2-table2}
report_table(make_individual_tree_table(rv2, tree = 2, condition_label = "name", digits = params$digits),
           paste(label_2, "Tree 2"))
```

### Comparison table

```{r rvine-tree2-compare}
report_table(make_compare_trees_table(rv1, rv2, tree = 2, label_a = label_1, label_b = label_2,
                                    condition_label = "name", digits = params$digits),
           paste(params$main_model_type, "Tree 2 comparison across clusters"))
```

```{r appendix-pagebreak, results="asis", echo=FALSE}
cat("\newpage")
```

# Appendix: CVine tree plots

```{r cvine-appendix, fig.height=10}
cvine_fits <- vine_fit_list$CVine
cv1 <- cvine_fits[[1]]
cv2 <- cvine_fits[[2]]

plot_individual_tree(cv1, tree = 1, node_label = params$node_label, condition_label = params$condition_label,
                     edge_label = "tau", min_abs_tau = params$min_abs_tau, digits = params$digits,
                     orientation = "radial", title = paste("CVine Cluster", cv1$cluster, "- Tree 1"))
plot_individual_tree(cv2, tree = 1, node_label = params$node_label, condition_label = params$condition_label,
                     edge_label = "tau", min_abs_tau = params$min_abs_tau, digits = params$digits,
                     orientation = "radial", title = paste("CVine Cluster", cv2$cluster, "- Tree 1"))
plot_individual_tree(cv1, tree = 2, node_label = params$node_label, condition_label = params$condition_label,
                     edge_label = "tau+cond", min_abs_tau = params$min_abs_tau, digits = params$digits,
                     orientation = "vertical", title = paste("CVine Cluster", cv1$cluster, "- Tree 2"))
plot_individual_tree(cv2, tree = 2, node_label = params$node_label, condition_label = params$condition_label,
                     edge_label = "tau+cond", min_abs_tau = params$min_abs_tau, digits = params$digits,
                     orientation = "vertical", title = paste("CVine Cluster", cv2$cluster, "- Tree 2"))
```
